---
title: "1_primary_responses_Solution"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
#source(here("code", "functions", "functions.R"))

```

# Objective 

In this document, we evaluate how long-lived IgA antibody titers in the respiratory mucosa interfere with the transport of specific IgA antibodies to the site of infection. 

# How to measure efficiency 

We might be interested in the effect of competition on a few different quantities: 

  1) The concentration of specific antibody that makes it to the lumen, $L_s$, over time, whose dynamics are described by: 

$$
\frac{dL_s}{dt} = rR_s - d_L L_s
$$
  
  2) The relative proportion of transport proteins carrying specific antibody out of the total number of transport proteins carrying antibody, given by:

$$
f_s = \frac{R_s}{R_s + R_{ns}}
$$

  3) The efficiency of transport of specific antibody relative to the amount of mucosal antibody, given by: 

$$
\epsilon_s = \frac{rR_s}{M_s}
$$

These might all be plotted as a function of time (x-axis) with various lines describing the dynamics given different $d_\mu$. 

However, we might also want to plot a heatmap of the effect of saturation on the response given changes in $d_\mu$ and strength of competition (either $K_M$ or $V_{max}$ changing). For these heatmaps, we will need a single value to report. My suggestion is either to report one of these values at the peak of the specific response ($t_{peak}$), or to plot the integral of one of these values up until the peak of the response. 

I will make that decision after seeing the results of the dynamics for each measurement! 

# Dynamics 

Below, we define the parameters required for the model `IgA_competition.model` in `code\functions\function.r`. This model is described in `code\simulations\1_primary_responses_Outline.qmd`. Importantly, we define a vector of $d_\mu$ to identify the effect of larger and larger concentration of total IgA on inhibitin the transport of newly generated IgA to a novel pathogen. 

```{r define-parameters}
### parameters for competition model 
# steady state nonspecific antibodies
c = #rate of influx of nonspecific IgA 
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years
d_mus = log(2)/t_halfs #half lives in days

# generation of new response
A0 = 1 #initial ab titer
Ap = 10^3 #peak antibody titer
tp = 5 #time of peak titer -- days

#will be calculated when iterating through d_mu
#r0 = (2*log(Ap/A0)/tp) - d_mu #initial/greatest exponential expansion rate
#alpha = (r0 - d_mu)/t_p # rate of decay of exponential growth rate

# transport 
k = #rate of antibody binding 
r = #rate of transport
RT = #total number (concentration) of receptors

# dynamics in the lumen 
t_half_L = 3 #half life of anitbodies in the lumen, in days
d_L = #decay rate in the lumen
```



