---
title: "1_primary_responses_Solution"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
source(here("code", "functions", "functions.R"))

```

# Objective 

In this document, we evaluate how long-lived IgA antibody titers in the respiratory mucosa interfere with the transport of specific IgA antibodies to the site of infection. 

# How to measure efficiency 

We might be interested in the effect of competition on a few different quantities: 

  1) The concentration of specific antibody that makes it to the lumen, $L_s$, over time, whose dynamics are described by: 

$$
\frac{dL_s}{dt} = rR_s - d_L L_s
$$
  
  2) The relative proportion of transport proteins carrying specific antibody out of the total number of transport proteins carrying antibody, given by:

$$
f_s = \frac{R_s}{R_s + R_{ns}}
$$

  3) The efficiency of transport of specific antibody relative to the amount of mucosal antibody, given by: 

$$
\epsilon_s = \frac{rR_s}{M_s}
$$

These might all be plotted as a function of time (x-axis) with various lines describing the dynamics given different $d_\mu$. 

However, we might also want to plot a heatmap of the effect of saturation on the response given changes in $d_\mu$ and strength of competition (either $K_M$ or $V_{max}$ changing). For these heatmaps, we will need a single value to report. My suggestion is either to report one of these values at the peak of the specific response ($t_{peak}$), or to plot the integral of one of these values up until the peak of the response. 

I will make that decision after seeing the results of the dynamics for each measurement! 

# Dynamics 

Below, we define the parameters required for the model `IgA_competition.model` in `code\functions\function.r`. This model is described in `code\simulations\1_primary_responses_Outline.qmd`. Importantly, we define a vector of $d_\mu$ to identify the effect of larger and larger concentration of total IgA on inhibitin the transport of newly generated IgA to a novel pathogen. 

```{r define-parameters-antibody-production}
### parameters for competition model 
# steady state nonspecific antibodies
c = 10^-2 #rate of influx of nonspecific IgA, Ab/time - comes from pb/dA - in mg/dl per day
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years
d_mus = log(2)/t_halfs #half lives in days

# generation of new response
Ap = (10^-3)*c/d_mus[length(d_mus)] #peak antibody titer will be 1/1000th of the highest concentration of ss. 
#initial ab titer
A0 = Ap/10^3 #1 1000th of the peak is where we start
tp = 5 #time of peak titer -- days

# will be calculated when iterating through d_mu
r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

#plot of change in r for specific cell production
for(d_mu in d_mus){
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate
  curve(r0 + alpha*x, from = 0, to = tp, add = T)
}

### plot Ms vs Mns over time
Mab_titers = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = c/d_mu # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  ### deSolve for specific ab over time
  
  # parms list
  params = list( # parameter list
  r0 = r0, 
  A0 = A0,
  Ap = Ap,
  alpha = alpha,
  tp = tp,
  d_mu = d_mu
  )

  # initial conditions
  Ms_init = A0 # starting antibody titer
  init_cond = c("Ms" = Ms_init)

  # numerical solver parameters
  Tmax = 30 # total days
  points_per_day = 24 # every hour

  # solve 
  soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), primary_response.model, parms = params)) %>% cbind(.,Mns, t_half)
}) %>% do.call(rbind, .) #concatenates all dfs from lapply list into a single df rowwise

### Plot
plot = ggplot() + geom_line(data = Mab_titers, aes(x = time, y = log10(Ms), group = factor(t_half), col = factor(t_half), linetype = "Ms")) +
  geom_line(data = Mab_titers, aes(x = time, y = log10(Mns), group = factor(t_half), col = factor(t_half), linetype = "Mns")) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("Ms" = "solid", "Mns" = "dashed")) +
  labs(x = "time, days", y = "[Ab], mg/dL", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "[Ms] and [Mns] over the course of infection") +
  theme_bw()
  

#print to viewer
print(plot)

```

```{r define-parameters-transport}
# transport 
k = #rate of antibody binding 
r = #rate of transport
RT = #total number (concentration) of receptors

# dynamics in the lumen 
t_half_L = 3 #half life of anitbodies in the lumen, in days
d_L = #decay rate in the lumen
```

```{r plot-steady-state}


```