---
title: "Simulations of specific IgA transport in the context of pre-existing immunity"
subtitle: "Simplified model: approximation, $M_{ns}$ constant"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
source(here("code", "functions", "functions.R"))

#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective 

In this document, we evaluate how long-lived IgA antibody titers in the respiratory mucosa interfere with the transport of specific IgA antibodies to the site of infection. We assume that the concentration of $M_{ns}$ changes negligibly over the course of infection, and assume that the steady state concentration of $M_{ns}$ is proportional to $\frac{c}{d_\mu}$. steady state concentration of non-specific antibodies explicitly considering the flux of IgA out of the mucosa. 

# Outline

1. Plot the generation of specific IgA, $M_s$ versus the steady state concentration of $M_{ns}$, $\frac{c}{d_\mu}$ over time. 
2. Show the MM plot of the flux of $M_s$ as a function of $M_s$, varying the half life of antibodies in the mucosa (via $d_\mu$)

- plot $K_M$ for the Michaelis-Menten kinetics 
- plot $A_p$, the peak of the specific antibody response

This will show us what regime of transport we are in over the course of infection, and where the half - saturation $[M_s]$ moves in the presence of a competitive inhibitor. 

3. Show the dynamics of $L_s$ over the course of infection (+ a week or two) for varying $d_\mu$

# Generation of $M_s$ compared to steady-state $M_{ns}$ given $d_\mu$

The generation of specific antibodies in the mucosa, $M_s$, follows the function: 

$$
\frac{dM_s}{dt} = r(t)M_s -k_1M_s(R_T(t) - R_s - R_{ns}) - d_\mu M_s
$$
where 

$$
r(t \leq t_{peak}) = r_0 - \alpha t \text{ and } r(t > t_{peak}) = 0
$$
and $\alpha = \frac{r_0 - d_{\mu}}{t_{peak}}$, $r_0 = \frac{2log(\frac{Ap}{A0})}{t_p} + d_{\mu}$

We assume that nonspecific antibodies, $M_{ns}$, are at steadys tate $\frac{c}{d_\mu}$ and change neglibily over the course of infection, such that: 

$$
\frac{dM_{ns}}{dt} = c - d_\mu M_{ns}
$$
for time until infection, 

$$
M_{ns}* = \frac{c}{d_\mu}
$$

is the steady state solution and the titer at the start of infection, and 

$$
\frac{dM_{ns}}{dt} = 0 
$$
during infection. 

We maintain binding of pIgRs, $R$, to these antibody types in a competitive manner: 

$$
\frac{dR_s}{dt} = k_1M_s(R_T(t)-R_s-R_{ns}) - k_2R_s
$$

$$
\frac{dR_{ns}}{dt} = k_1M_{ns}(R_T(t)-R_s-R_{ns}) - k_2R_{ns}
$$

And the lumenal titer $L$ of each type increases with transport and decreases with degredation in the lumen:

$$
\frac{dL_s}{dt} = k_2R_s - d_L L_s
$$

$$
\frac{dL_{ns}}{dt} = k_2R_{ns} - d_L L_{ns}
$$

Notice that currently I am not changing the concentration of either the specific nor the nonspecific antibodies according to what gets transported into the lumen. 

I guess I would change the specific more than the nonspecific, but also, if we choose the nonspecific to be equal to the ss titer of nonspecific with the median halflife, that would be silly. So for now, neither changes, and only the receptors change. 

## Dynamics in mucosa during infection 

This gives rise to the following dynamics in the mucosa over the course of infection: 

```{r define-parameters}
# steady state nonspecific antibodies
c = 1*10^5 #rate of influx of nonspecific IgA, Ab/time - comes from pb/dA - in mg/dl per day
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years
d_mus = log(2)/t_halfs #vector of decay rates, half lives in days
names(d_mus) = t_halfs
Mns_vec = c/d_mus #vecotr of approximate ss solutions
names(Mns_vec) = t_halfs

# generation of new response
Ap = Mns_vec["90"] #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = 1 # initial specific ab concentration
tp = 5 #time of peak titer -- days

# will be calculated when iterating through d_mu
#r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
#alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

# transport
Vmax = Ap # order of magnitude of peak response # we might vary this later
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
k1 = k2/min(Mns_vec) #k2/min(Mns_vec) #rate of antibody binding such that KM is order of magnitude of min of the Mns
KM = k2/k1 #checking KM

# dynamics in the lumen 
t_half_L = 3 #half life of antibodies in the lumen, in days
d_L = log(2)/t_half_L #decay rate in the lumen

```

```{r plot-Ms-and-Mns}
### plot Ms vs Mns over time
Mab_titers = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = c/d_mu # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  # set time vector parms
  Tmax = 30 # total days
  points_per_day = 24 # every hour
  time = seq(0, Tmax, length.out = Tmax*points_per_day) #generate time vector

  ### apply to calculate
  soln = data.frame("time" = time, "Ms" = sapply(time, function(t){
    if(t<=tp){A0*exp((r0 - d_mu)*t - (alpha/2)*t^2)}
    else {Ap*exp(-d_mu*(t-tp))}
  }))
  

  # solve 
  soln = soln %>% cbind(.,Mns, t_half)
}) %>% do.call(rbind, .) #concatenates all dfs from lapply list into a single df rowwise

### Plot
plot = ggplot() + geom_line(data = Mab_titers, aes(x = time, y = log10(Ms), group = factor(t_half), col = factor(t_half), linetype = "Ms")) +
  geom_line(data = Mab_titers, aes(x = time, y = log10(Mns), group = factor(t_half), col = factor(t_half), linetype = "Mns")) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("Ms" = "solid", "Mns" = "dashed")) +
  labs(x = "time, days", y = "log10[Ab], mg/dL", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "[Ms] and [Mns] over the course of infection") +
  theme_bw()
  

#print to viewer
print(plot)

```

## Michaelis-Menten flux curves by $M_{ns}$

Below, we plot the flux of specific antibody into the lumen, $\phi_s$ as a function of the concnetration of specific antibody in the mucosa, $M_s$, assuming michaelis-menten type kinetics (**check how much our approximations affect the assumptions**):

$$
\phi_s = \frac{V_{max}M_s}{K_M + M_{ns} + M_s}
$$
Where $V_{max} = k_2 R_T$, $K_M = \frac{k_2}{k_1}$, and reminder that $M_{ns} \approx \frac{c}{d_\mu}$.

In effect, the half-saturation point of $[M_s]$ without competition, given by $K_M$, shifts to the right by $M_{ns}$.

```{r define-MM}
#define MM function
MM_function = function(M_s, M_ns) {(Vmax*M_s)/(KM+M_ns+M_s)}
```

```{r plot-MM}
#set concentrations of Ms for x axis
  concentrations = 10^seq(1, 10, by = 0.1) #range of concentrations M_s such that we span much less than and much greater than Ap

### plot flux vs [Ms] given each Mns 
flux_df = lapply(t_halfs, function(t_half){
  # set d_mu and Mns 
  d_mu = log(2)/t_half
  Mns = c/d_mu
  
  #calculate flux
  flux = sapply(concentrations, function(Ms){
   
   return(MM_function(M_s = Ms, M_ns = Mns))
  }) %>% cbind("Ms" = concentrations, "flux" = ., "t_half" = t_half) %>% as.data.frame() 
}) %>% do.call(rbind, .) %>% as.data.frame() 

## plot
plot = ggplot() + geom_line(data = flux_df, aes(x = (Ms), y = flux, group = factor(t_half), col = factor(t_half))) +
  geom_hline(yintercept = Vmax, linetype = "dashed") +
  geom_vline(xintercept = (k2/k1), color = "red") +
  geom_vline(xintercept = (Ap), linetype= "dashed") +
geom_line(
  aes(x = concentrations, y = sapply(concentrations, function(Ms){Vmax*Ms/(KM+Ms)})),
  inherit.aes = FALSE,
  color = "red"
) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = my_colors) +
  labs(x = "[Ms]", y = "flux [Ms]", color = "Mucosal Ab Half Life, days", title = "Competitive inhibition of specific antibody transport by nonspecific antibodies") + 
  theme_bw()

# print to viewer
print(plot)
```

The horizontal dashed line shows the greatest possible transport rate, $V_{max}$; the red curve shows the MM kinetics given $[M_s]$ only; the vertical red line shows the 50% saturation concentration given $[M_s]$ only, $K_M$; and I have added a vertical dashed line to show the peak concentration of $M_s$ generated over an infection.

# Dynamics over infection

Because this is a simplified model, I want to plot both the output of the MM function over the course of generation the specific response given a constant SS concentration of nonspecific antibodies, as well as the output of an ODE model that explicitly models the binding and transport of specific and non-specific antibodies. Hopefully theyll look very similar. 

```{r model-function}
# michaelis menten kinetics - desolve ode model
simplified_competition.model <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        Ms = y[1]
        Mns = y[2]
        Rs = y[3]
        Rns = y[4]
        Ls = y[5]
        Lns = y[6]
        
        # make empty variables for the derivatives
        dMs = NaN
        dMns = NaN
        dRs = NaN
        dRns = NaN
        dLs = NaN
        dLns = NaN
        
        #####################################
        # define time dependent functions   #
        #####################################
        # exponential growth rate of Ms at time t
        r = function(t) {if(t < tp) {r0 - alpha*t}
          else 0} #define time-varying r(t) - increase that slows and is then dominated by decay rate
        
        # calculate Re at time t
        Re = RT - Rs - Rns # total number of receptors - those bound to specific ab - those bound to nonspecific ab
        
        #####################################
        #	 calculate the derivatives
        dMs = r(t)*Ms - d_mu*Ms  # specific Ab production and decay
        dMns = 0
        dRs = k1*Ms*Re - k2*Rs
        dRns = k1*Re*Mns - k2*Rns
        dLs = k2*Rs - d_L*Ls
        dLns = k2*Rns - d_L*Lns
        
        # output the derivatives
        dy=c(dMs, dMns, dRs, dRns, dLs, dLns) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} 


```




```{r ODE-model-plots}
### numerical simulation parms
Tmax_ss = 10000 # to get ss L_ns
points_per_day_ss = 1

Tmax = 30
points_per_day = 24 

### deSolve soltions
ODE_solns = lapply(t_halfs, function(t_half){ # half lives in days
  ### set parms 
    d_mu = log(2)/t_half #decay rate per day
    Mns_init = c/d_mu
    
  # As list
  params = list( #setting A
    # nonspecific ss
    d_mu = d_mu, # plasma cell decay rate per day
    c = c, # total antibody production rate per day
    #generation of new responses
    Ap = Ap, # peak antibody titer
    A0 = A0, # intial antibody titer
    tp = tp, # time of peak response, days
    r0 = (2*log(Ap/A0)/tp) + d_mu, #initial/greatest exponential expansion rate
    alpha = (((2*log(Ap/A0)/tp) + d_mu) - d_mu)/tp, # rate of decay of exponential growth rate
    # transport
    k2 = k2, # trancytosis rate
    RT = RT, # total number of receptors
    k1 = k1, # ab binding rate
    # lumen
    d_L = d_L #decay rate in the lumen
  )
  
  ############ solve for steady state concentrations ###################
  init_cond = c("Ms" = 0, "Mns" = Mns_init, "Rs" = 0, "Rns" = 0, "Ls" = 0, "Lns" = 0)

  # ode solve for steady state without specific response (to account for transport of nonspecific to lumen)
  init_cond = as.data.frame(ode(y = init_cond, times = seq(0, Tmax_ss, length.out = Tmax_ss*points_per_day_ss), simplified_competition.model, parms = params)) %>% tail(.,1) %>% select(!time) %>% unlist(.)
  
  ############## course of infection ######################
  init_cond["Ms"] = A0 # set specific antibody to increase
  
  soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), simplified_competition.model, parms = params)) %>% cbind(., t_half) 
  
  return(soln)
}) %>% do.call(rbind, .) %>% pivot_longer(cols = !c(time, t_half),
                                          names_to = "population",
                                          values_to = "concentration") %>% as.data.frame()


#plot spectific responses in the mucosa
plot = ggplot() +
  geom_line(data = ODE_solns %>% filter(population %in% c("Mns", "Ms")), aes(x = time, y = log10(concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = factor(population))) +
  scale_y_continuous(limits = c(-1,NA)) +
  scale_color_manual(values = my_colors) +
  theme_bw()

# print to viewer
print(plot)

#plot spectific responses in the lumen
plot = ggplot() +
  geom_line(data = ODE_solns %>% filter(population %in% c("Lns", "Ls")), aes(x = time, y = log10(concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = factor(population))) +
  scale_y_continuous(limits = c(-1,NA)) +
  scale_color_manual(values = my_colors) +
  theme_bw()

# print to viewer
print(plot)

# Plotting flux over time to compare 
plot = ggplot() +
  geom_line(data = ODE_solns %>% filter(population %in% c("Rns", "Rs")), aes(x = time, y = log10(k2*concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = factor(population))) +
  scale_y_continuous(limits = c(-1,NA)) +
  labs(y = "flux, k2*Ri") +
  scale_color_manual(values = my_colors) +
  theme_bw()

#print to viewer
print(plot)

```

```{r MM-plots}
### plot Ms vs Mns over time
MM_soln = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = c/d_mu # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  # set time vector parms
  Tmax = 30 # total days
  points_per_day = 24 # every hour
  time = seq(0, Tmax, length.out = Tmax*points_per_day) #generate time vector

  ### apply to calculate Ms titers over time
  Ms_soln = data.frame("time" = time, "t_half" = t_half, "Ms" = sapply(time, function(t){
    if(t<=tp){A0*exp((r0 - d_mu)*t - (alpha/2)*t^2)}
    else {Ap*exp(-d_mu*(t-tp))}
  }), "Mns" = Mns)
  
  ### apply to calculate Ls flux over time
  ### 
  Ms_soln$phi_s = sapply(Ms_soln$Ms, function(Ms){MM_function(M_s = Ms, M_ns = Mns)})
  Ms_soln$phi_ns = sapply(Ms_soln$Ms, function(Ms){(Vmax*Mns)/(KM+Ms+Mns)})
  
  return(Ms_soln)
}) %>% do.call(rbind, .)

### Plot
plot = ggplot() + geom_line(data = MM_soln, aes(x = time, y = log10(phi_s), group = factor(t_half), col = factor(t_half), linetype = "phi_s")) +
  geom_line(data = MM_soln, aes(x = time, y = log10(phi_ns), group = factor(t_half), col = factor(t_half), linetype = "phi_ns")) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("phi_s" = "solid", "phi_ns" = "dashed")) +
  labs(x = "time, days", y = "log10[R-Ab], mg/dL, over time", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "flux of s and ns into the lumen over the course of infection") +
  theme_bw()
  

#print to viewer
print(plot)

```


And if we want to see how similar these are, we can plot them together: 

```{r plot-MM-ODE-together}
plot = ggplot() +
  geom_line(data = ODE_solns %>% filter(population %in% c("Rns", "Rs")), aes(x = time, y = log10(k2*concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = "ODE")) +
  geom_line(data = MM_soln, aes(x = time, y = log10(phi_s), group = factor(t_half), col = factor(t_half), linetype = "MM")) +
  geom_line(data = MM_soln, aes(x = time, y = log10(phi_ns), group = factor(t_half), col = factor(t_half), linetype = "MM")) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("ODE" = "solid", "MM" = "dashed")) +
  labs(x = "time, days", y = "log10[R-Ab], mg/dL, over time", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "flux of s and ns into the lumen over the course of infection") +
  theme_bw() +
  scale_y_continuous(limits = c(-1,NA)) +
  theme_bw()

#print to viewer
print(plot)
```

Aww they're basically the same! All that running to stay in one place :) heck, I'm doing a PhD!

And I bet in this parameter regime the full ode will give the same result. Aw shall we try? 

# Compare to the ODE including transport in steady state

```{r compare-approximation-full}
### parameters for competition model 
### define initial conditions and parms
# initial conditions
  Ms_init = 0 # starting antibody titer
  Mns_init = 0
  Rs_init = 0
  Rns_init = 0
  Ls_init = 0
  Lns_init = 0

### numerical simulation parms
Tmax_ss = 10000
points_per_day_ss = 1

Tmax = 30
points_per_day = 24 

### deSolve soltions
full_solns = lapply(t_halfs, function(t_half){ # half lives in days

  ############ solve for steady state concentrations ###################
  init_cond = c("Ms" = 0, "Mns" = Mns_init, "Rs" = 0, "Rns" = 0, "Ls" = 0, "Lns" = 0)
  ### set parms 
    d_mu = log(2)/t_half #decay rate per day
    
  # As list
  params = list( #setting A
    # nonspecific ss
    d_mu = d_mu, # plasma cell decay rate per day
    c = c, # total antibody production rate per day
    #generation of new responses
    Ap = Ap, # peak antibody titer
    A0 = A0, # intial antibody titer
    tp = tp, # time of peak response, days
    r0 = (2*log(Ap/A0)/tp) + d_mu, #initial/greatest exponential expansion rate
    alpha = (((2*log(Ap/A0)/tp) + d_mu) - d_mu)/tp, # rate of decay of exponential growth rate
    # transport
    k2 = k2, # trancytosis rate
    RT = RT, # total number of receptors
    k1 = k1, # ab binding rate
    # lumen
    d_L = d_L #decay rate in the lumen
  )
  
  # ode solve for steady state without specific response (to account for transport of nonspecific to lumen)
  init_cond = as.data.frame(ode(y = init_cond, times = seq(0, Tmax_ss, length.out = Tmax_ss*points_per_day_ss), IgA_competition.model, parms = params)) %>% tail(.,1) %>% select(!time) %>% unlist(.)
  
  ############## course of infection ######################
  init_cond["Ms"] = A0 # set specific antibody to increase
  
  soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), IgA_competition.model, parms = params)) %>% cbind(., t_half) 
  
  return(soln)
}) %>% do.call(rbind, .) %>% pivot_longer(cols = !c(time, t_half),
                                          names_to = "population",
                                          values_to = "concentration") %>% as.data.frame()



```

```{r plot-all}
plot = ggplot() +
  geom_line(data = ODE_solns %>% filter(population %in% c("Rns", "Rs")), aes(x = time, y = log10(k2*concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = "simple_ODE")) +
   geom_line(data = full_solns %>% filter(population %in% c("Rns", "Rs")), aes(x = time, y = log10(k2*concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = "full_ODE")) +
  geom_line(data = MM_soln, aes(x = time, y = log10(phi_s), group = factor(t_half), col = factor(t_half), linetype = "MM")) +
  geom_line(data = MM_soln, aes(x = time, y = log10(phi_ns), group = factor(t_half), col = factor(t_half), linetype = "MM")) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("simple_ODE" = "solid", "MM" = "dashed", "full_ODE" = "dotted")) +
  labs(x = "time, days", y = "log10[R-Ab], mg/dL, over time", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "flux of s and ns into the lumen over the course of infection") +
  theme_bw() +
  scale_y_continuous(limits = c(-1,NA)) +
  theme_bw()

#print to viewer
print(plot)
```


Looks like the full ode predicts more flux of specific antibody than the approximations for high concentrations of $M_{ns}$. So if we want to be conservative about the effect, we might go with the full model?