---
title: "Michaelis-Mentin Approximation of Antibody Transport"
subtitle: "To what extent does competition from previously generated, non-specific antibodies interfer with newly generated responses"
format: pdf
editor: visual
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
source(here("code", "functions", "functions.R"))

here("code", "functions", "functions.R")

#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective

# Objective 

In this document, we evaluate how long-lived IgA antibody titers remaining from previous infections of the respiratory mucosa interfere with the transport of newly generated, specific IgA antibodies to the site of infection. We assume that the concentration of $M_{ns}$ changes negligibly over the course of infection, and assume that the steady state concentration of $M_{ns}$ is proportional to $\frac{c}{d_\mu}$. We use a Michaelis-Menten approximation to calculate the flux of specific IgA across the mucosal membrane, and treat nonspecific antibodies as a competitive inhibitor of the transport process.

# Outline 

I will review the kinetics inferred by the Michaelis-Mentin approximation before presenting the results of competitive inhibition by non-specific immunity during a specific response. 

We are modelling the factors that contribute to the strength of competition between newly generated, pathogen specific IgA ($M_s$) and nonspecific IgA remaining from previous infections ($M_{ns}$) in the mucosa for transport across the epithelium and to the cite of infection, the Lumen. 

We are interested in how the half life of IgA titers in the mucosa - thus, the steady-state concentration of non-specific IgA in the mucosa - affects the magnitude and efficiency of newly-generated specific IgA responses. We therefore compare the effect of varying IgA half-life ($t_{1/2}$) on these two quantities to the effect of changing the rate of transport of antibodies across the epithelium ($k_2$) and effect of increases the concentration of transport proteins ($R_T$) on the mucosal surface.  

The results in this document are organized as follows: 

1) How do $t_{1/2}$, $k_2$, and $R_T$ effect the flux of antibody into the lumen over the course of infection (VAR vs Time)

2) How do $t_{1/2}$ and $k_2$ affect i) the peak flux of the antibody and ii) the efficiency of the antibody transport to the lumen? 

3) How do $t_{1/2}$ and $R_T$ affect  i) the peak flux of the antibody and ii) the efficiency of the antibody transport to the lumen? 

# Michaelis-Mentin Kinetics

Below, we define our parameters and vary the $t_{1/2}$ of mucosal antibody titers to observe how pre-existing antibodies increase the concentration of specific IgA at which point the transport of specific antibodies is half-saturated:

```{r define-parameters-vary-thalf}
# transport
Vmax = 10^3 # order of magnitude of peak response # we might vary this later
RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
k1 = k2/10^3 #k2/min(Mns_vec) #rate of antibody binding such that KM is on the order of 10^3
KM = k2/k1 #checking KM

# steady state nonspecific antibodies
c = 1*10^5 #rate of influx of nonspecific IgA, Ab/time - comes from pb/dA - in mg/dl per day
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years :) 69 
d_mus = log(2)/t_halfs #half lives in days
  names(d_mus) = t_halfs
Mns_vec = c/d_mus #vecotr of approximate ss solutions
  names(Mns_vec) = t_halfs

# generation of new response
Ap = Mns_vec[round(length(t_halfs)/2)] #peak antibody titer will be the same as Mns for 90 days - the middle value of all Mns

```

Below, we plot how the flux of specific antibody for a given concentration of $M_s$ in the mucosa changes as a function of the half-life of non-specific antibodies. In essence, the effective amount of $M_s$ required to reach half-maximal transport will increase by the concentration of non-specific antibody conferred by the increasing half-life (the effective $K_M = K_M + \hat M_{ns}$.) 

```{r plot-MM-kinetics}
#set concentrations of Ms for x axis
concentrations = 10^seq(1, 10, by = 0.1) #range of concentrations M_s such that we span much less than and much greater than Ap

### plot flux vs [Ms] given each Mns 
flux_df = lapply(t_halfs, function(t_half){
  # set d_mu and Mns 
  d_mu = log(2)/t_half
  Mns = c/d_mu
  
  #calculate flux using MM function in functions.R
  flux = sapply(concentrations, function(Ms){
  return(MM_function(Vmax = Vmax, KM = KM, M_s = Ms, M_ns = Mns))
  }) %>% 
    cbind("Ms" = concentrations, "flux" = ., "t_half" = t_half) %>% # add the concentrations of Ms correspond to each flux as a column, and label all observations with the thalf used to calculate
    as.data.frame() #format as a dataframe
}) %>% 
  do.call(rbind, .) %>% as.data.frame() #combine all dataframes to make one long-from data frame with a column keeping track of thalf used for calculating each observation

## plot
plot = ggplot() + geom_line(data = flux_df, aes(x = (Ms), y = flux, group = factor(t_half), col = factor(t_half))) + #plot flux by thalf
  geom_hline(yintercept = Vmax, linetype = "dashed") + #add vmax as horizontal line
  geom_vline(xintercept = KM, color = "red") + #add half-saturation point in absence of competition as vertical line
  geom_vline(xintercept = Ap, linetype= "dashed") + #add vertical line showing peak titer generated in mucosa
geom_line(
  aes(x = concentrations, y = sapply(concentrations, function(Ms){MM_function(Vmax = Vmax, KM = KM, M_s = Ms, M_ns = 0)})),
  inherit.aes = FALSE,
  color = "red"
) + #add line showing how flux depends on Ms in the absence of competition
  scale_x_continuous(trans = "log10") + #x axis as log scale
  scale_color_manual(values = my_colors) + #set colors to my color scheme
  labs(x = "[Ms]", y = "flux [Ms]", color = "Mucosal Ab Half Life, days", title = "Competitive inhibition of specific antibody transport by nonspecific antibodies") +  #label axes 
  theme_bw() #changes background color and grid lines

# print to viewer
print(plot)
```

The horizontal dashed line shows the greatest possible transport rate, $V_{max}$; the red curve shows the MM kinetics given $[M_s]$ only; the vertical red line shows the 50% saturation concentration given $[M_s]$ only, $K_M$; and I have added a vertical dashed line to show the peak concentration of $M_s$ generated over an infection.

Notice that I have chosen paramerters such that, in the absence of competition, the maximum flux is reached by the the maximum concentration of $M_s$ generated during a response. As nonspecific antibody is added, however, the flux achieved at the maximum specific antibody concentration drops. This is the core of the competitive effect. 


# Factors that affect the flux of antibody into the lumen 

The competitive affect can be reduced by increasing the rate of transport of antibody across the membrane, as well as increasing the maximum transport rate, $V_{max}$. Both of these will increase the flux of specific antibody as well as nonspecific antibody across. Below, I plot how both of these factors change the flux of $M_s$ in the presence of nonspecific antibody. I set the half life = 90 days for this example. 

```{r define-parameters-vary-k2}
# transport
RT = 10^3/(1/(1/24)) #total number (concentration) of receptors, Vmax = k2RT
k2_vec = 1/(c(0.25,0.5,0.75,1,2,3)/24)
Vmax = k2_vec*RT #Vmax increases as well as KM when k2 increases
k1 = k1 # use k1 defined earlier
KM_vec = k2_vec/k1

# steady state nonspecific antibodies
c = 1*10^5 #rate of influx of nonspecific IgA, Ab/time - comes from pb/dA - in mg/dl per day
t_half = 90 #half life of mucosal ab in days  
  d_mu = log(2)/t_half #half life in days
  Mns = c/d_mu #approximate ss solution
 
# generation of new response
Ap = Mns_vec[round(length(t_halfs)/2)] #peak antibody titer will be the same as Mns for 90 days - the middle value of all Mns

```

```{r plot-MM-kinetics-k2}
#set concentrations of Ms for x axis
concentrations = 10^seq(1, 10, by = 0.1) #range of concentrations M_s such that we span much less than and much greater than Ap

### plot flux vs [Ms] given each Mns 
flux_df = lapply(k2_vec, function(k2){
  # set vmax and KM
  Vmax = k2*RT
  KM = k2/k1
  
  #calculate flux using MM function in functions.R
  flux = sapply(concentrations, function(Ms){
  return(MM_function(Vmax = Vmax, KM = KM, M_s = Ms, M_ns = Mns))
  }) %>% 
    cbind("Ms" = concentrations, "flux" = ., "k2" = k2) %>% # add the concentrations of Ms correspond to each flux as a column, and label all observations with the thalf used to calculate
    as.data.frame() #format as a dataframe
}) %>% 
  do.call(rbind, .) %>% as.data.frame() #combine all dataframes to make one long-from data frame with a column keeping track of thalf used for calculating each observation

## plot
plot = ggplot() + geom_line(data = flux_df, aes(x = (Ms), y = flux, group = factor(k2), col = factor(k2))) + #plot flux by thalf
  geom_vline(xintercept = KM, color = "red") + #add half-saturation point in absence of competition as vertical line
  geom_vline(xintercept = Ap, linetype= "dashed") + #add vertical line showing peak titer generated in mucosa
geom_line(
  aes(x = concentrations, y = sapply(concentrations, function(Ms){MM_function(Vmax = Vmax, KM = KM, M_s = Ms, M_ns = 0)})),
  inherit.aes = FALSE,
  color = "red"
) + #add line showing how flux depends on Ms in the absence of competition
  scale_x_continuous(trans = "log10") + #x axis as log scale
  scale_color_manual(values = my_colors) + #set colors to my color scheme
  labs(x = "[Ms]", y = "flux [Ms]", color = "k2, days-1", title = "Competitive inhibition of specific antibody transport by nonspecific antibodies") +  #label axes 
  theme_bw() #changes background color and grid lines

# print to viewer
print(plot)
```

