---
title: "Simulations of specific IgA transport in the context of pre-existing immunity"
subtitle: "Accounting for flux of $M_{ns}$ into the lumen when calculating Steady State $[M_{ns}]$"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
source(here("code", "functions", "functions.R"))

#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective 

In this document, we evaluate how long-lived IgA antibody titers in the respiratory mucosa interfere with the transport of specific IgA antibodies to the site of infection. We explore the steady state concentration of non-specific antibodies explicitly considering the flux of IgA out of the mucosa. 

# How to measure efficiency 

We might be interested in the effect of competition on a few different quantities: 

  1) The concentration of specific antibody that makes it to the lumen, $L_s$, over time, whose dynamics are described by: 

$$
\frac{dL_s}{dt} = k_2R_s - d_L L_s
$$
  
  2) The relative proportion of transport proteins carrying specific antibody out of the total number of transport proteins carrying antibody, given by:

$$
f_s = \frac{R_s}{R_s + R_{ns}}
$$

  3) The efficiency of transport of specific antibody relative to the amount of mucosal antibody, given by: 

$$
\epsilon_s = \frac{k_2R_s}{M_s}
$$

These might all be plotted as a function of time (x-axis) with various lines describing the dynamics given different $d_\mu$. 

However, we might also want to plot a heatmap of the effect of saturation on the response given changes in $d_\mu$ and strength of competition (either $K_M$ or $V_{max}$ changing). For these heatmaps, we will need a single value to report. My suggestion is either to report one of these values at the peak of the specific response ($t_{peak}$), or to plot the integral of one of these values up until the peak of the response. 

I will make that decision after seeing the results of the dynamics for each measurement! 

# Parameters 

Below, we define the parameters required for the model `IgA_competition.model` in `code\functions\function.r`. This model is described in `code\simulations\1_primary_responses_Outline.qmd`. Importantly, we define a vector of $d_\mu$ to identify the effect of larger and larger concentration of total IgA on inhibitin the transport of newly generated IgA to a novel pathogen. 

## Antibody concentrations 

First, we define how we vary the concentration of nonspecific antibody $M_{ns}$ with respect to the dynamics of specific antibody $M_s$ increase and decay during infection.

Importantly, the steady state concentration of $M_{ns}$ is not simply the steady state of $\frac{dM_{ns}}{dt} = c - d_\mu$ which gives $\frac{c`}{d_\mu}$, because antibody is constantly being transported into the lumen. Instead, the equilibrium concentration of $M_{ns}$ should be derived as the positive solution to the steady state of the following system of ODEs:

$$
\frac{dM_{ns}}{dt} = c - k_1 (R_T - R_{ns}) M_{ns} - d_\mu M_{ns}
$$

and

$$
\frac{dR_{ns}}{dt} = k_1(R_T - R_{ns})M_{ns} - k_2R_{ns}
$$

solving for the steady state gives: 

$$
M_{ns}^{*} = \frac{-\left(d_{\mu}k_{2}+k_{1}k_{2}R_{T}-ck_{1}\right)+\sqrt{\left(d_{\mu}k_{2}+k_{1}k_{2}R_{T}-ck_{1}\right)^{2}+4d_{\mu}k_{1}ck_{2}}}{2d_{\mu}k_{1}}
$$


Below, we plot the dependence of $M_{ns}^{*}$ on the decay rate ($d_\{mu}$). 

First, we define the function to calculate $M_{ns}^{*}$:

```{r steady-state-nonsepcific}
#define steady state concentration of mns
ss_Mns = function(d_mu) {
  num = -(d_mu*k2 + k1*k2*RT - c*k1) +
    sqrt((d_mu*k2 + k1*k2*RT - c*k1)^2 + 4*d_mu*k1*c*k2)

  den = 2*d_mu*k1

  return(num / den)
}
```

```{r define-parameters-transport}
# transport
Vmax = 10^3 # order of magnitude of peak response # we might vary this later
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
k1 = k2/10^3 #k2/min(Mns_vec) #rate of antibody binding such that KM is order of magnitude of min of the Mns
KM = k2/k1 #checking KM

# dynamics in the lumen 
t_half_L = 3 #half life of antibodies in the lumen, in days
d_L = log(2)/t_half_L #decay rate in the lumen

# steady state nonspecific antibodies
c = 1*10^5 #rate of influx of nonspecific IgA, Ab/time - comes from pb/dA - in mg/dl per day
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years
d_mus = log(2)/t_halfs #half lives in days
names(d_mus) = t_halfs

```

```{r ss_by_d_mu}
### generate ss by d_mu
Mns_vec = sapply(d_mus, function(d_mu){ss_Mns(d_mu = d_mu)}) #calculate mns for each dmu
signif(Mns_vec,1) 

##plot
plot = ggplot() + 
  geom_point(aes(x = log10(t_halfs), y = log10(Mns_vec))) +
  geom_line(aes(x = log10(t_halfs), y = log10(Mns_vec))) +
  theme_classic()

#print to viewer 
print(plot)

### we might also want to see how this depends on vmax...
```


We would like to plot the specific antibody response with respect to the steady state concentration of non-specific antibodies. Here, the generation and decay of specific antibody in the mucosa (not considering transport) looks like:

$$
M_s(t) = A_0 \exp{(r_0 - d_\mu) t - \frac{\alpha}{2}t^2}
$$

```{r parameters-specific-responses}
# generation of new response
Ap = Mns_vec["90"] #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = 1 # initial specific ab concentration
tp = 5 #time of peak titer -- days

# will be calculated when iterating through d_mu
#r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
#alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

```

```{r plot-specific-responses}
### plot Ms vs Mns over time
Mab_titers = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = ss_Mns(d_mu = d_mu) # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  ### deSolve for specific ab over time
  
  # parms list
  params = list( # parameter list
  r0 = r0, 
  A0 = A0,
  Ap = Ap,
  alpha = alpha,
  tp = tp,
  d_mu = d_mu
  )

  # initial conditions
  Ms_init = A0 # starting antibody titer
  init_cond = c("Ms" = Ms_init)

  # numerical solver parameters
  Tmax = 30 # total days
  points_per_day = 24 # every hour

  # solve 
  soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), primary_response.model, parms = params)) %>% cbind(.,Mns, t_half)
}) %>% do.call(rbind, .) #concatenates all dfs from lapply list into a single df rowwise

### Plot
plot = ggplot() + geom_line(data = Mab_titers, aes(x = time, y = log10(Ms), group = factor(t_half), col = factor(t_half), linetype = "Ms")) +
  geom_line(data = Mab_titers, aes(x = time, y = log10(Mns), group = factor(t_half), col = factor(t_half), linetype = "Mns")) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("Ms" = "solid", "Mns" = "dashed")) +
  labs(x = "time, days", y = "log10[Ab], mg/dL", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "[Ms] and [Mns] over the course of infection") +
  theme_bw()
  

#print to viewer
print(plot)
```

I also explicitly plot the integral of the ode used to generate the specific response, but i've hidden that for now because its identical (which is what we wouldve hoped for!) 
```{r check-specific-response, include: FALSE}
### plot Ms vs Mns over time
Mab_titers = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = ss_Mns(d_mu = d_mu) # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  # set time vector parms
  Tmax = 30 # total days
  points_per_day = 24 # every hour
  time = seq(0, Tmax, length.out = Tmax*points_per_day) #generate time vector

  ### apply to calculate
  soln = data.frame("time" = time, "Ms" = sapply(time, function(t){
    if(t<=tp){A0*exp((r0 - d_mu)*t - (alpha/2)*t^2)}
    else {Ap*exp(-d_mu*(t-tp))}
  }))
  

  # solve 
  soln = soln %>% cbind(.,Mns, t_half)
}) %>% do.call(rbind, .) #concatenates all dfs from lapply list into a single df rowwise

### Plot
plot = ggplot() + geom_line(data = Mab_titers, aes(x = time, y = log10(Ms), group = factor(t_half), col = factor(t_half), linetype = "Ms")) +
  geom_line(data = Mab_titers, aes(x = time, y = log10(Mns), group = factor(t_half), col = factor(t_half), linetype = "Mns")) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("Ms" = "solid", "Mns" = "dashed")) +
  labs(x = "time, days", y = "log10[Ab], mg/dL", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "[Ms] and [Mns] over the course of infection") +
  theme_bw()
  

#print to viewer
print(plot)

```

When we solve numerically and plot the concentration of $M_s$, it will be smaller, due to flux outside. But we wanted the peak of the response generated to match the order of magnitude of the steady state nonspecific titer. 

# Effect of competition

Below, we plot the effect of pre-existing antibody titers on the flux of specific antibodies across the lumen. We do so fist by showing how the curve of flux by the concentration os $M_s$ changes in the presence of increase $M_{ns}$.

```{r plot-Ms-flux-curve-by-Mns}
#set concentrations of Ms for x axis
  concentrations = 10^seq(-3, 10, by = 0.1)

### plot flux vs [Ms] given each Mns 
flux_df = lapply(t_halfs, function(t_half){
  # set d_mu and Mns 
  d_mu = log(2)/t_half
  Mns = ss_Mns(d_mu = d_mu)
  
  #calculate flux
  flux = sapply(concentrations, function(Ms){
    Vmax*Ms/((k2/k1) + Mns + Ms)
  }) %>% cbind("Ms" = concentrations, "flux" = ., "t_half" = t_half) %>% as.data.frame() 
}) %>% do.call(rbind, .) %>% as.data.frame() 

## plot
plot = ggplot() + geom_line(data = flux_df, aes(x = (Ms), y = flux, group = factor(t_half), col = factor(t_half))) +
  geom_hline(yintercept = Vmax, linetype = "dashed") +
  geom_vline(xintercept = (k2/k1), linetype = "dashed") +
  geom_vline(xintercept = (Ap), color = "red") +
geom_line(
  aes(x = concentrations, y = sapply(concentrations, function(Ms){Vmax*Ms/(KM+Ms)})),
  inherit.aes = FALSE,
  color = "red"
) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = my_colors) +
  labs(x = "[Ms]", y = "flux [Ms]", color = "Mucosal Ab Half Life, days", title = "Competitive inhibition of specific antibody transport by nonspecific antibodies") + 
  theme_bw()

# print to viewer
print(plot)
```

The horizontal dashed line shows the greatest possible transport rate, $V_{max}$, the vertical dashed line shows the 50% saturation concentration given $[M_s]$ only, or $K_M$, and I have added a red vertical line to show the peak concentration of $M_s$ generated over an infection.


# Dynamics as a function of mucosal antibody half-life

We would like to show how high concentrations of nonspecific antibody in the mucosa interfere with the transport of specific antibody across the mucosa and into the lumen, the site of infection. 

## Concentration of specific antibody, $L_s$ in the lumen over time

Here, we numerically solve for the concentration of nonspecific and specific antibody in the mucosa over time: 

```{r primary_response_mucosal_compartment}
### parameters for competition model 
## define half-lifes
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years

### define initial conditions and parms
# initial conditions
  Ms_init = 0 # starting antibody titer
  Mns_init = 0
  Rs_init = 0
  Rns_init = 0
  Ls_init = 0
  Lns_init = 0

### numerical simulation parms
Tmax_ss = 10000
points_per_day_ss = 1

Tmax = 30
points_per_day = 24 

### deSolve soltions
ss_solns = lapply(t_halfs, function(t_half){ # half lives in days

  ############ solve for steady state concentrations ###################
  init_cond = c("Ms" = 0, "Mns" = Mns_init, "Rs" = 0, "Rns" = 0, "Ls" = 0, "Lns" = Lns_init)
  ### set parms 
    d_mu = log(2)/t_half #decay rate per day
    
  # As list
  params = list( #setting A
    # nonspecific ss
    d_mu = d_mu, # plasma cell decay rate per day
    c = c, # total antibody production rate per day
    #generation of new responses
    Ap = Ap, # peak antibody titer
    A0 = A0, # intial antibody titer
    tp = tp, # time of peak response, days
    r0 = (2*log(Ap/A0)/tp) + d_mu, #initial/greatest exponential expansion rate
    alpha = (((2*log(Ap/A0)/tp) + d_mu) - d_mu)/tp, # rate of decay of exponential growth rate
    # transport
    k2 = k2, # trancytosis rate
    RT = RT, # total number of receptors
    k1 = k1, # ab binding rate
    # lumen
    d_L = d_L #decay rate in the lumen
  )
  
  # ode solve for steady state without specific response (to account for transport of nonspecific to lumen)
  init_cond = as.data.frame(ode(y = init_cond, times = seq(0, Tmax_ss, length.out = Tmax_ss*points_per_day_ss), IgA_competition.model, parms = params)) %>% tail(.,1) %>% select(!time) %>% unlist(.)
  
  ############## course of infection ######################
  init_cond["Ms"] = A0 # set specific antibody to increase
  
  soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), IgA_competition.model, parms = params)) %>% cbind(., t_half) 
  
  return(soln)
}) %>% do.call(rbind, .) %>% pivot_longer(cols = !c(time, t_half),
                                          names_to = "population",
                                          values_to = "concentration") %>% as.data.frame()


#plot spectific responses in the lumen
plot = ggplot() +
  geom_line(data = ss_solns %>% filter(population %in% c("Mns", "Ms")), aes(x = time, y = log10(concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = factor(population))) +
  scale_y_continuous(limits = c(-1,NA)) +
  scale_color_manual(values = my_colors) +
  theme_bw()

# print to viewer
print(plot)



```

And now we want to see what happens in the lumen:

```{r primary_response_lumen_dynamics}
#plot spectific responses in the lumen
plot = ggplot() +
  geom_line(data = ss_solns %>% filter(population %in% c("Lns", "Ls")), aes(x = time, y = log10(concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = population)) +
  scale_y_continuous(limits = c(-1,NA)) +
  scale_color_manual(values = my_colors) +
  theme_bw()

# print to viewer
print(plot)

```

# Parking on the downhill


