---
title: "Evolutionary Constraints on the Longevity of IgA Antibody Titers"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
source(here("code", "functions", "functions.R"))


```

# Objective

The goal of this document is to udnerstand the extent to which antibodies remaining from prior respiratory infections inhibit the transport of newly generated, specific IgA during infection. 

# Setting parameters

```{r set-parms}
# biological parameters for competition simulation 
t.half = c(1/12, 6/12, 1, 2, 5, 10)*365 #half lives in years converted to days
## for steady state Mns
  c = 0.5 #constant rate of IgA generated through time, per day
  d_mus = -log(1/2)/(t.half*365) #decay rate of antibody in the mucosa (1/days)
## for newly generated Ms
  A0 = 100
  r = log(10^2)/14 #exponential growth of new population
  t_p = 7 #1/2 time to peak (days)
## for transport and lumen concentration
  d_l = 1/10 #1/days
  K = 100
  Vmax = 10

```

# Running Sim

## Calculating steady state Mucosal IgA

```{r transport-vs-d_l}
# Calculate ss concentrations of Mns
M_ns.vec = sapply(d_mus, function(d_mu) {M_ns(c = c, d = d_mu)})

# plot 
plot(x = t.half/365, y = log10(M_ns.vec), xlab = "Half life of mucosal antibody titers, Years", ylab = "Log10 Total M_ns", )

#create df to store d_mu, M_ns, and t.half 
M_ns.df = data.frame( "t.half" = t.half,  #half life
                      "d_mu" = d_mus, #corresponding decay rate
                      "M_ns" = M_ns.vec) #corresponding ss antibody titer

```

## Specific Antibodies generated during infection 

```{r plot-specific-antibodies}
# mucosal specific IgA during infection (no decay)
times = seq(0, 2*t_p , by = 1/24) #every hour for t < t_max days
M_s.infection = sapply(times, function(t) {M_s(A0 = A0, t_p = t_p, r = r, t = t)}) #calculate dynamics of specific IgA on the inside

plot(x = times, y = M_s.infection, type = "l", xlab = "Day post infection", ylab = "Specific Antibody Inside, M_s")

```

## Lumenal IgA as a function of internal mucosal decay rate, $d_\mu$

```{r setup-Competion}
## set up parms list for deSolve
params = list(
  c = c, #constant rate of IgA generated through time, per day, A0/tau (in days)
  d_mu = NaN, #decay rate of antibody in the mucosa (1/days)
# for newly generated Ms
  A0 = A0,
  r = r, #exponential growth of new population
  t_p = t_p, #1/2 time to peak (days)
# for transport and lumen concentration
  d_l = d_l, #1/days
  K = K, #half max concentration
  Vmax = Vmax #max transport rate
)

## deSolve parameters
# simulation parameters 
Tmax   =   30               #20       # max time in days - 1 month
points.per.day = 24                # output points per year

# initial conditions
init.cond = c(L_s = 0) # initial mucosal iga

```


```{r competition-by-dmu}
#solve for lumen specific antibody dynamics
competition.sims = lapply(1:nrow(M_ns.df), function(i){
  # extract ss M_ns information
  t.half = M_ns.df[i, "t.half"] #extract half life
  d_mu = M_ns.df[i, "d_mu"] #extract decay rate
  M_ns.val = M_ns.df[i, "M_ns"] #extract ss M_ns value
  
  # substitute d_mu into parms
  params$d_mu = d_mu
  
  # deSolve
  soln = as.data.frame(ode(y = init.cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), IgA_competition, parms = params))
  
  # add cols to include t.half, d_mu, and M_ns.val 
  soln = cbind(soln, "t.half" = t.half, "d_mu" = d_mu, "M_ns" = M_ns.val)
})

```

```{r plot-lumen-dynamics-by-dmu}
#combine all dataframes to one
solns = do.call(rbind, competition.sims) %>% as.data.frame()

#plot
plot = ggplot() + geom_line(data = solns, aes(x = time, y = L_s, group = factor(signif(t.half/365,2)), col = factor(signif(t.half/365, 2)))) +
  theme_classic() +
  labs(x = "days post infection", y = "IgA in lumen", col = "decay rate internal, years")

#print to viewer
print(plot)


```

# Varying d_mu and K

Now, we investigate how the total IgA transported by day 14 changes as a function of $K$ and $d_\mu$:

```{r params}
#define sequence of Ks and t.halfs
Ks = seq(1, 10000, by = 100)
t.halfs = seq(1/12,10, by = 0.1)*365 #half lives in years converted to days

## set up parms list for deSolve
params = list(
  c = c, #constant rate of IgA generated through time, per day, A0/tau (in days)
  d_mu = NaN, #decay rate of antibody in the mucosa (1/days)
# for newly generated Ms
  A0 = A0,
  r = r, #exponential growth of new population
  t_p = t_p, #1/2 time to peak (days)
# for transport and lumen concentration
  d_l = d_l, #1/days
  K = NaN, #half max concentration
  Vmax = Vmax #max transport rate
)

## deSolve parameters
# simulation parameters 
Tmax   =   30               #20       # max time in days - 1 month
points.per.day = 24                # output points per year

# initial conditions
init.cond = c(L_s = 0) # initial mucosal iga

```

```{r competition-heatmap}
#calculate total transported by day 14 by K and t.half
total.Ls = sapply(Ks, function(K){ #rows
  sapply(t.halfs, function(t.half){ #columns
    #set parms
    params$d_mu = -log(1/2)/(t.half*365)
    params$K = K
    
    #competition
    # deSolve
  soln = as.data.frame(ode(y = init.cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), IgA_competition, parms = params))
  
  #calculate sum of antibodies to day 14 
  total.Ls.transported = sum(soln %>% filter(time < 14) %>% pull(L_s))
  })
})
```

```{r plot-ks-vs-dmus}
# matrix
filled.contour(x = t.halfs/365, y = Ks, z=log10(total.Ls), lwd = 2, xlab = "decay rate internal, years", ylab = "Michaelis-menten K", main = "Integral of IgA in Lumen by day 14") #draw figure

```
