---
title: "Evolutionary Constraints on the Longevity of IgA Antibody Titers"
subtitle: "Simulations Part 1: constraint on the longevity of antibody titers in the respiratory mucosa"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#install.packages("RColorBrewer")
library(RColorBrewer)

#load functions
source(here("code", "functions", "functions.R"))

#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective 

The goal of this document is to understand the extent to which antibodies remaining from prior respiratory infections inhibit the transport of newly generated, specific IgA during infection. 

# Outline

We will generate two figures: 

1) The flux of $M_s$ to the lumen as a function of $[M_s]$ on the x-axis; steady-state $[M_{ns}]$ or, in other words, $d_\mu$, changing line colors; and a low and high competition setting by high and low $V_{max}$ by (two) plots. 

To generate our high and low competition settings, we want to consider the transcytosis rate, $k_{cat}$, and $[pIgR]$ concentrations separately. $k_cat$ will remain the same in the high and low competition settings, but $[pIgR]$ will change to simulate high/low abundance of transport proteins. 

We also want to set our $K$ such that the peak of the specific antibody response is about $K$.

# Setting parameters

```{r set-parms}
# biological parameters for competition simulation 
t_half = c(1/52, 1/12, 1, 10, 100)*365 #half lives of antibody titers in years converted to days
## for steady state Mns
  c = 1 #constant rate of IgA generated through time, per day
  d_mus = -log(1/2)/(t_half) #decay rate of antibody in the mucosa (1/days)
  
  ss_Mns = (signif( (c/d_mus), 2)) # 1,2,3,4,5!

## for newly generated Ms
  A0 = 10^5
  r =  log10(A0)/14 #exponential growth of new population
  t_p = 7 #1/2 time to peak (days)

  # dynamics of antibody responses over course of infection
  #specific antibodies during infection
  M_s = function(A0, t_p, r, d_mu, t) {(A0/(1+exp(-r*(t-t_p))))} #sigmoidal growth, A0 is max antibody, t_p is half the time to peak, r is initial growt

  t = seq(0,3*t_p, by = 1/24)

  m_trajectories = sapply(d_mus, function(d_mu){
    ms = sapply(t, function(time) {M_s(A0 = A0, t_p = t_p, r = r, t = time)})
    }) %>% cbind(t, .) %>% as.data.frame()
  colnames(m_trajectories) = c("time", t_half)
    
  ## for transport and lumen concentration
  d_l = 1/10 #1/days
  Ks = c(c/d_mus[3], (c/d_mus[3])*10^4) #middle 
  V_max = 10^6 #catalytic rate and number of enzymes, per day (every 24 hours)

```

Justifying parms *HERE*

## Plots 

Below, we calculate the rate of transport of $M_s$ across the mucosal epithelium as a function of log10 ($M_s$) (x-axis), half life of mucosal antibodies as seperate lines, and two plots, a high competition (low $V_{max}$) and low competition (high $V_{max}$) setting. 

```{r calculate-transport}
# function for calculating flux
Specific_flux = function(M_s, M_ns, K, V_max){M_s*V_max/(K+M_s+M_ns)}
# set parms
M_s.vec = seq(0,(max(m_trajectories)), by = 100) 

# calculate curves 
M_s_transport = lapply(Ks, function(K){
  Flux_df = sapply(ss_Mns, function(Mns){
    sapply(M_s.vec, function(Ms){
      Specific_flux(M_s = Ms, M_ns = Mns, K = K, V_max = V_max)
    })
  }) %>% as.data.frame() %>% cbind(M_s.vec, ., K)
  colnames(Flux_df) = c("M_s", t_half, "K")

  Flux_df = Flux_df %>% pivot_longer(
    cols = !c(M_s, K),
    names_to = "t_half",
    values_to = "flux"
  )
})

M_s_transport = do.call(rbind, M_s_transport) %>% as.data.frame()

plot = ggplot() + 
    geom_line(data = M_s_transport, aes(x = (M_s), y = log10(flux), col = t_half, linetype = as.factor(K))) +
    theme_bw() +
  labs(x = "Specific Antibody, Inside", y = "Specific Antibody Flux", linetype = "Concentration of \nhalf-saturated transport", color = "Antibody Half Life, Interior") +
  scale_linetype_manual(values = c("solid", "dashed"), labels = signif(Ks, 2)) +
    scale_color_manual(values = my_colors, labels = signif(t_half, 2))

#print to viewer
print(plot)

```

What if I plot as a ratio of what gets over vs what is left? Flux with or without competition? Proportion specific to nonsepcific? 

This is absolutely what we expect but not quite the right rational... I need to measure something dimensionless. 

```{r calculate-reduction-transport}
# function for calculating flux
Specific_flux = function(M_s, M_ns, K, V_max){M_s*V_max/(K+M_s+M_ns)}
# set parms
M_s.vec = seq(0,(max(m_trajectories)), by = 100) 

# calculate curves 
M_s_transport = lapply(Ks, function(K){
  Flux_df = sapply(ss_Mns, function(Mns){
    sapply(M_s.vec, function(Ms){
      competition = Specific_flux(M_s = Ms, M_ns = Mns, K = K, V_max = V_max)
      no_competition = Specific_flux(M_s = Ms, M_ns = 0, K = K, V_max = V_max)
      return(competition/no_competition)
    })
  }) %>% as.data.frame() %>% cbind(M_s.vec, ., K)
  colnames(Flux_df) = c("M_s", t_half, "K")

  Flux_df = Flux_df %>% pivot_longer(
    cols = !c(M_s, K),
    names_to = "t_half",
    values_to = "prop_flux"
  )
})

M_s_transport = do.call(rbind, M_s_transport) %>% as.data.frame()

plot = ggplot() + 
    geom_line(data = M_s_transport, aes(x = (M_s), y = log10(prop_flux), col = t_half, linetype = as.factor(K))) +
    theme_bw() +
  labs(x = "Specific Antibody, Inside", y = "Prop. Specific Antibody Flux", linetype = "Concentration of \nhalf-saturated transport", color = "Antibody Half Life, Interior") +
  scale_linetype_manual(values = c("solid", "dashed"), labels = signif(Ks, 2)) +
    scale_color_manual(values = my_colors, labels = signif(t_half, 2))

#print to viewer
print(plot)

```


#########################

# Setting parameters

```{r set-parms}
# biological parameters for competition simulation 
t_half = c(1/52, 1/12, 1, 10, 100)*365 #half lives of antibody titers in years converted to days
## for steady state Mns
  c = 1 #constant rate of IgA generated through time, per day
  d_mus = -log(1/2)/(t_half) #decay rate of antibody in the mucosa (1/days)
  
  ss_Mns = (signif( (c/d_mus), 2)) # 1,2,3,4,5!

## for newly generated Ms
  A0 = 10^5
  r =  log10(A0)/14 #exponential growth of new population
  t_p = 7 #1/2 time to peak (days)

  # dynamics of antibody responses over course of infection
  #specific antibodies during infection
  M_s = function(A0, t_p, r, d_mu, t) {(A0/(1+exp(-r*(t-t_p))))} #sigmoidal growth, A0 is max antibody, t_p is half the time to peak, r is initial growt

  t = seq(0,3*t_p, by = 1/24)

  m_trajectories = sapply(d_mus, function(d_mu){
    ms = sapply(t, function(time) {M_s(A0 = A0, t_p = t_p, r = r, t = time)})
    }) %>% cbind(t, .) %>% as.data.frame()
  colnames(m_trajectories) = c("time", t_half)
    
  ## for transport and lumen concentration
  d_l = 1/10 #1/days
  K = c/d_mus[3] #middle 
  V_max.vec = c(10, 10^6) #catalytic rate and number of enzymes, per day (every 24 hours)

```

Justifying parms *HERE*

## Plots 

Below, we calculate the rate of transport of $M_s$ across the mucosal epithelium as a function of log10 ($M_s$) (x-axis), half life of mucosal antibodies as seperate lines, and two plots, a high competition (low $V_{max}$) and low competition (high $V_{max}$) setting. 

```{r calculate-transport}
# function for calculating flux
Specific_flux = function(M_s, M_ns, K, V_max){M_s*V_max/(K+M_s+M_ns)}
# set parms
M_s.vec = seq(0,(max(m_trajectories)), by = 100) 

# calculate curves 
M_s_transport = lapply(V_max.vec, function(V_max){
  Flux_df = sapply(ss_Mns, function(Mns){
    sapply(M_s.vec, function(Ms){
      Specific_flux(M_s = Ms, M_ns = Mns, K = K, V_max = V_max)
    })
  }) %>% as.data.frame() %>% cbind(M_s.vec, ., V_max)
  colnames(Flux_df) = c("M_s", t_half, "V_max")

  Flux_df = Flux_df %>% pivot_longer(
    cols = !c(M_s, V_max),
    names_to = "t_half",
    values_to = "flux"
  )
})

M_s_transport = do.call(rbind, M_s_transport) %>% as.data.frame()

plot = ggplot() + 
    geom_line(data = M_s_transport, aes(x = (M_s), y = log10(flux), col = t_half, linetype = as.factor(V_max))) +
    theme_bw() +
  labs(x = "Specific Antibody, Inside", y = "Specific Antibody Flux", linetype = "V_max", color = "Antibody Half Life, Interior") +
  scale_linetype_manual(values = c("solid", "dashed"), labels = signif(V_max, 2)) +
    scale_color_manual(values = my_colors, labels = signif(t_half, 2))

#print to viewer
print(plot)

```

What if I plot as a ratio of what gets over vs what is left? Flux with or without competition? Proportion specific to nonsepcific? 

This is absolutely what we expect but not quite the right rational... I need to measure something dimensionless. 

```{r calculate-reduction-transport}
# function for calculating flux
Specific_flux = function(M_s, M_ns, K, V_max){M_s*V_max/(K+M_s+M_ns)}
# set parms
M_s.vec = seq(0,(max(m_trajectories)), by = 100) 

# calculate curves 
M_s_transport = lapply(V_max.vec, function(V_max){
  Flux_df = sapply(ss_Mns, function(Mns){
    sapply(M_s.vec, function(Ms){
      competition = Specific_flux(M_s = Ms, M_ns = Mns, K = K, V_max = V_max)
      no_competition = Specific_flux(M_s = Ms, M_ns = 0, K = K, V_max = V_max)
      return(competition/no_competition)
    })
  }) %>% as.data.frame() %>% cbind(M_s.vec, ., V_max)
  colnames(Flux_df) = c("M_s", t_half, "V_max")

  Flux_df = Flux_df %>% pivot_longer(
    cols = !c(M_s, V_max),
    names_to = "t_half",
    values_to = "prop_flux"
  )
})

M_s_transport = do.call(rbind, M_s_transport) %>% as.data.frame()

plot = ggplot() + 
    geom_line(data = M_s_transport, aes(x = (M_s), y = log10(prop_flux), col = t_half, linetype = as.factor(V_max))) +
    theme_bw() +
  labs(x = "Specific Antibody, Inside", y = "Specific Antibody Flux", linetype = "V_max", color = "Antibody Half Life, Interior") +
  scale_linetype_manual(values = c("solid", "dashed"), labels = signif(V_max.vec, 2)) +
    scale_color_manual(values = my_colors, labels = signif(t_half, 2))

#print to viewer
print(plot)

```

