---
title: "Evolutionary Constraints on the Longevity of IgA Antibody Titers:"
subtitle: "Outline Part 2: Inducible plasma cells rapidly recall IgA responses and control initial growth of viral infection"
author: "Cora Hirst, Susi Linderman, Carl Davis, Hasan Ahmed, Rustom Antia..."
output: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
#source(here("code", "functions", "functions.R"))

#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective

The goal of this document is to explore a biological mechanism of IgA recall responses in the respiratory tract - namely that, rather than constitutively secreting antibody, localized IgA plasma cells are inducible - through ODE models. 

# Outline 

I plan to formulate 2 models that both result in inducibility: 

1) Once plasma cells are generated, they differentiate into "quiescent" plasma cells at some rate. On recall responses, they are recalled at some rate. 

2) Once plasma cells are generated, they only secrete antibody in the presence of antigen. When antigen disappears, secretion stops. 

I think the first will be more gradated/biological for a model with bianry "infection/non-infection". But, if we allow the rate of recruitment (1) or the rate of secretion (2) to be dependent on the density of antigen, then I'm not entirely sure how different the two models will be!

# Model 1 - Plasma cell differentiation

Below, we define an ode model in which there are 3 populations of cells: B cells, secretory plasma cells, and quiescent plasma cells. We include a fourth variable - "Virus" - which is either 0 (no infection present) or 1 (infection present). 

When $V = 0$, each cell population decays at some rate, and secretory plasma cells differentiate into quiescent plasma cells. 

When $V = 1$, B cells proliferate and differentiate into secretory plasma cells. Secretory plasma cells differentiate into quiescent plasma cells at some rate. Quiescent plasma cells are recruited to differentiate into secretory plasma cells at some rate. Cell populations continue to decay according to their given death rate. 

```{r ODE-Model-1}
# Antibody concentrations in the respiratory mucosa #
secondary_response.model <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        B = y[1]         # B cells 
        P_s = y[2]       # Antibody secreting cells (specific)
        P_q = y[3]       # quiescent antibody secreting cells (specific)
        M_s = y[4]       # specific Ab in mucosa
        L_s = y[5]       # specific Ab in lumen
        
        
        # make empty variables for the derivatives
        dB = 0      # change in specific Ab in the lumen
        dP_s = 0    # change in specific ASCs
        dP_q = 0    # change in quiescent ASCs
        dM_s = 0    # change in specific antibodies in mucosa
        dL_s = 0    # change in specific antibodies in the lumen
        
        #####################################
        # define V function 
        V.function = function(t_infection, t) {
          if(t>=min(t_infection) & t <= max(t_infection)){
            V = 1}
          else(V = 0)
        return(V)}
        # set V
        V = V.function(t_infection, t)
        #####################################
        ## calculate the derivatives
        # cell populations
        dB = r_b*(1-diff_frac)*B*V - diff_frac*B*V -d_B*B #B cells (memory and naive together)
        dP_s = diff_frac*B*V - quiessence_k*P_s - d_mu*P_s + induction_k*P_q*V # specific ASCs
        dP_q = quiessence_k*P_s - induction_k*P_q*V - d_q*P_q # quiescent sepecific ASCs
        # Specific antibodies        
        dM_s = rho*P_s - (Vmax/(K + M_s + M_ns(c = c, d = d_mu)))*M_s - d_Ab*M_s #mucosal
        dL_s = (Vmax/(K + M_s + M_ns(c = c, d = d_mu)))*M_s - d_l*L_s #lumenal
        
        # output the derivatives
        dy=c(dB, dP_s, dP_q, dM_s, dL_s) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} 

```

S

