---
title: "Evolutionary Constraints on the Longevity of IgA Antibody Titers:"
subtitle: "Outline Part 2: Inducible plasma cells rapidly recall IgA responses and control initial growth of viral infection"
author: "Cora Hirst, Susi Linderman, Carl Davis, Hasan Ahmed, Rustom Antia..."
output: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
#source(here("code", "functions", "functions.R"))

#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective

The goal of this document is to explore a biological mechanism of IgA recall responses in the respiratory tract - namely that, rather than constitutively secreting antibody, localized IgA plasma cells are inducible - through ODE models. 

# Outline 

I plan to formulate 2 models that both result in inducibility: 

1) Once plasma cells are generated, they differentiate into "quiescent" plasma cells at some rate. On recall responses, they are recalled at some rate. 

2) Once plasma cells are generated, they only secrete antibody in the presence of antigen. When antigen disappears, secretion stops. 

I think the first will be more gradated/biological for a model with bianry "infection/non-infection". But, if we allow the rate of recruitment (1) or the rate of secretion (2) to be dependent on the density of antigen, then I'm not entirely sure how different the two models will be!

# Model 1 - Plasma cell differentiation

Below, we define an ode model in which there are 3 populations of cells: B cells, secretory plasma cells, and quiescent plasma cells. We include a fourth variable - "Virus" - which is either 0 (no infection present) or 1 (infection present). 

When $V = 0$, each cell population decays at some rate, and secretory plasma cells differentiate into quiescent plasma cells. 

When $V = 1$, B cells proliferate and differentiate into secretory plasma cells. Secretory plasma cells differentiate into quiescent plasma cells at some rate. Quiescent plasma cells are recruited to differentiate into secretory plasma cells at some rate. Cell populations continue to decay according to their given death rate. 

```{r define-ODE-Model-1}
# Antibody concentrations in the respiratory mucosa #
secondary_response.model1 <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        B = y[1]         # B cells 
        P_s = y[2]       # Antibody secreting cells (specific)
        P_q = y[3]       # quiescent antibody secreting cells (specific)
        M_s = y[4]       # specific Ab in mucosa
        
        # make empty variables for the derivatives
        dB = 0      # change in specific Ab in the lumen
        dP_s = 0    # change in specific ASCs
        dP_q = 0    # change in quiescent ASCs
        dM_s = 0    # change in specific antibodies in mucosa
        
        #####################################
        # define V function 
        V.function = function(t_infection, t) {
          if(t>=min(t_infection) & t <= max(t_infection)){
            V = 1}
          else(V = 0)
        return(V)}
        # set V
        V = V.function(t_infection, t)
        #####################################
        ## calculate the derivatives
        # cell populations
        dB = r*(1-f)*V*B - d_B*B #proliferation only when virus present, some fraction f differentiates to P_s
        dP_s = r*f*V*B + s*V*P_q - d_s*P_s - q*P_s #differentiation from B cells or P_q only when virus present
        dP_q = q*P_s - d_q*P_q - s*V*P_q #P_s always differentiates to P_q with some rate q.
        
        # Specific antibodies        
        dM_s = a*P_s - d_a*M_s #P_s secretes antibody
        
        # output the derivatives
        dy=c(dB, dP_s, dP_q, dM_s) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} # we will need to think about how to set M_ns if we want to track specific IgA in the lumen.

```

```{r set-parms-ODE-Model-1}
# death rates
d_B = log(2)/(90) # B cell death rate, x days half life
d_s = log(2)/90 # secretory cell death rate #don't consider for now
d_q = log(2)/(365*2) # quiescent cell death rate long lived like bm
  
# proliferation and differentiation
r = 0.3 # b cell proliferation rate - doubling time is 1/2 day
f = 1/3 # fraction of division that produce secretory cells
q = log(2)/30 # rate of secretory cells to quiescence #should equal d_mu
s = 0.01 # rate of quiescent cell recruitment to secretion #should be very fast, like a percentage per hour

# antibody dynamics
a = 10 # rate of IgA secretion by secretory cells # not dealing with density but total number now
d_a = log(2)/3 # IgA molecular decay rate, 3 days half life

# infection timing
t_infection = c(0,14) #time range of antigen presence (V = 1 when t in t_infection)
```

```{r solve-ODE-Model-1-first-infection}
# parameters list for deSolve
params = list(
  d_B = d_B, # B cell death rate
  d_s = d_s, # secretory cell death rate
  d_q = d_q, # quiescent cell death rate
  d_a = d_a, # IgA molecular decay rate
    
  # proliferation and differentiation
  r = r, # b cell proliferation rate
  f = f, # fraction of division that produce secretory cells
  q = q, # rate of secretory cells to quiescence
  s = s, # rate of quiescent cell recruitment to secretion
  
  # antibody dynamics
  a = a, # rate of IgA secretion by secretory cells
  d_a = d_a, # IgA molecular decay rate in mucosa
  
  #infection timing
  t_infection = t_infection #time range of antigen presence (V = 1 when t in t_infection)
)

# simulation Time parameters 
Tmax   = 365*2                         # max time in days
points.per.day = 24              # output points per day

#initial conditions
init_cond = c("B" = 10, "P_s" = 0, "P_q" = 0, "M_s" = 0)

# solve 
soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model1, parms = params)) 

# long form ode 
soln_long = soln %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
   coord_cartesian(ylim = c(0,3)) +
   labs(x = "days",, y = bquote("log"[10] ~ "number"), title = "primary infection, inducible response") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)
```

```{r plot-ODE-Model-1-secondary-infection}
# set initial conditions
init_cond = soln %>% select(!(time)) %>% slice(nrow(soln)) %>% unlist()

# simulation Time parameters 
Tmax   = 30                         # max time in days
points.per.day = 24              # output points per day

# solve 
soln_secondary = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model1, parms = params)) 



# long form ode 
soln2_long = soln_secondary %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
   coord_cartesian(ylim = c(0,3)) +
  labs(x = "days",, y = bquote("log"[10] ~ "number"), title = "secondary infection, inducible response") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)
```

compare to if $d_\mu$ was actual death rate

```{r ODE-model-no-inducible}
# Antibody concentrations in the respiratory mucosa #
secondary_response.model_notinducible <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        B = y[1]         # B cells 
        P_s = y[2]       # Antibody secreting cells (specific)
        M_s = y[3]       # specific Ab in mucosa
        
        # make empty variables for the derivatives
        dB = 0      # change in specific Ab in the lumen
        dP_s = 0    # change in specific ASCs
        dM_s = 0    # change in specific antibodies in mucosa
        
        #####################################
        # define V function 
        V.function = function(t_infection, t) {
          if(t>=min(t_infection) & t <= max(t_infection)){
            V = 1}
          else(V = 0)
        return(V)}
        # set V
        V = V.function(t_infection, t)
        #####################################
        ## calculate the derivatives
        # cell populations
        dB = r*(1-f)*V*B - d_B*B #proliferation only when virus present, some fraction f differentiates to P_s
        dP_s = r*f*V*B - d_s*P_s #differentiation from B cells or P_q only when virus present
        
        # Specific antibodies        
        dM_s = a*P_s - d_a*M_s #P_s secretes antibody
        
        # output the derivatives
        dy=c(dB, dP_s, dM_s) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} # we will need to think about how to set M_ns if we want to track specific IgA in the lumen.
```

```{r set-parms-ODE-Model-1-noinducible}
# death rates
d_B = log(2)/(90) # B cell death rate, 2 years half life
d_s = log(2)/30 # rate of secretory cells to death #should equal d_mu
 
# proliferation and differentiation
r = 0.3 # b cell proliferation rate - doubling time is 1 day
f = 1/3 # fraction of division that produce secretory cells

# antibody dynamics
a = 10 # rate of IgA secretion by secretory cells # not dealing with density but total number now
d_a = log(2)/3 # IgA molecular decay rate, 3 days half life

# infection timing
t_infection = c(0,14) #time range of antigen presence (V = 1 when t in t_infection)
```

```{r solve-ODE-Model-1-first-infection-noinducible}
# parameters list for deSolve
params = list(
  d_B = d_B, # B cell death rate
  d_s = d_s, # secretory cell death rate

    
  # proliferation and differentiation
  r = r, # b cell proliferation rate
  f = f, # fraction of division that produce secretory cells

  # antibody dynamics
  a = a, # rate of IgA secretion by secretory cells
  d_a = d_a, # IgA molecular decay rate in mucosa
  
  #infection timing
  t_infection = t_infection #time range of antigen presence (V = 1 when t in t_infection)
)

# simulation Time parameters 
Tmax   = 365*2                         # max time in days
points.per.day = 24              # output points per day

#initial conditions
init_cond = c("B" = 10, "P_s" = 0, "M_s" = 0)

# solve 
soln_notinducible = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model_notinducible, parms = params)) 

# long form ode 
soln_notinducible_long = soln_notinducible %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln_notinducible_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln_notinducible_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
   coord_cartesian(ylim = c(0,3)) +
   labs(x = "days", y = bquote("log"[10] ~ "number"), title = "primary infection, not inducible response") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)

```

```{r plot-ODE-Model-1-secondary-infection-noinducible}
# set initial conditions
init_cond = soln_notinducible %>% select(!(time)) %>% slice(nrow(soln_notinducible)) %>% unlist()

# simulation Time parameters 
Tmax   = 30                         # max time in days
points.per.day = 24              # output points per day

# solve 
soln_notinducible_secondary = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model_notinducible, parms = params)) 



# long form ode 
soln_notinducible2_long = soln_notinducible_secondary %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln_notinducible2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln_notinducible2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
   labs(x = "days", y = bquote("log"[10] ~ "number"), title = "secondary infection, not inducible response") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)
```

plotting $M_s$ together: 

```{r plot-together}
plot = ggplot() + geom_line(data = soln2_long %>% filter(population == "M_s"), aes(x = time, y = (number), linetype = "inducible"), lwd = 1.5) + 
  geom_line(data = soln_notinducible2_long %>% filter(population == "M_s"), aes(x = time, y = (number), linetype = "not inducible"), lwd = 1.5) +
  scale_linetype_manual(values = c("inducible" = "solid", "not inducible" = "dashed")) +
  labs(x = "days", y = bquote("number"), title = "secondary infection, antibody response") +
  theme_bw() #removes grey background and changes aesthetics

#print to viewer
print(plot)


```

not as impressive on a log scale, but how much difference does it make on a linear scale, which might matter more during the early phase of a virus infection? How to show this, if it makes any sense? 

```{r compare-linearly}
plot = ggplot() + geom_line(data = soln2_long %>% filter(population == "M_s"), aes(x = time, y = (number), linetype = "inducible"), lwd = 1.5) + 
  geom_line(data = soln_notinducible2_long %>% filter(population == "M_s"), aes(x = time, y = (number), linetype = "not inducible"), lwd = 1.5) +
  scale_linetype_manual(values = c("inducible" = "solid", "not inducible" = "dashed")) +
  labs(x = "days", y = bquote("number"), title = "secondary infection, antibody response") +
  theme_bw() #removes grey background and changes aesthetics

#print to viewer
print(plot)


```



# Model 2 - Plasma cell tunable secretion

Below, we define an ode model in which there are 2 populations of cells: B cells and plasma cells. We include a third variable - "Virus" - which is either 0 (no infection present) or 1 (infection present). 

When $V = 0$, each cell population decays at some rate, and plasma cells do not secrete antibody. 

When $V = 1$, B cells proliferate and differentiate into plasma cells. Plasma cells secrete antibody. Cell populations continue to decay according to their given death rate. 

```{r define-ODE-Model-2, evaluate = F}
# Antibody concentrations in the respiratory mucosa #
secondary_response.model2 <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        B = y[1]         # B cells 
        P = y[2]       # Antibody secreting cells (specific)
        M_s = y[3]       # specific Ab in mucosa
        
        # make empty variables for the derivatives
        dB = 0      # change in specific Ab in the lumen
        dP = 0    # change in specific ASCs
        dM_s = 0    # change in specific antibodies in mucosa
        
        #####################################
        # define V function 
        V.function = function(t_infection, t) {
          if(t>=min(t_infection) & t <= max(t_infection)){
            V = 1}
          else(V = 0)
        return(V)}
        # set V
        V = V.function(t_infection, t)
        #####################################
        ## calculate the derivatives
        # cell populations
        dB = r*(1-f)*V*B - d_B*B #proliferation only when virus present, some fraction f differentiates to P
        dP = r*f*V*B - d_p*P  #differentiation from B cells only when virus present
        
        # Specific antibodies        
        dM_s = a*V*P - d_a*M_s #P secretes antibody only in the presence of virus
        
        # output the derivatives
        dy=c(dB, dP, dM_s) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} # we will need to think about how to set M_ns if we want to track specific IgA in the lumen.

```

```{r set-parms-ODE-Model-2, evaluate = F}
# death rates
d_B = log(2)/(365*50) # B cell death rate, 50 years half life # B cell death rate
d_p = log(2)/(365*50) # plasma cell death rate
d_a = log(2)/3 # IgA molecular decay rate, 3 days half life

  
# proliferation and differentiation
r = log(2)/1 # b cell proliferation rate - doubling time is 1/2 day
f = .2 # fraction of division that produce secretory cells
  
# antibody dynamics
a = 10^6/10^3 # rate of IgA secretion by secretory cells # not dealing with density but total number now
d_a = log(2)/3 # IgA molecular decay rate, 3 days half life

# infection timing
t_infection = c(0,14) #time range of antigen presence (V = 1 when t in t_infection)
```

```{r solve-ODE-Model-2, evaluate = F}
# parameters list for deSolve
params = list(
  d_B = d_B, # B cell death rate
  d_p = d_p, # plasma cell death rate
  d_a = d_a, # IgA molecular decay rate
    
  # proliferation and differentiation
  r = r, # b cell proliferation rate
  f = f, # fraction of division that produce secretory cells
  
  # antibody dynamics
  a = a, # rate of IgA secretion by secretory cells
  d_a = d_a, # IgA molecular decay rate in mucosa
  
  #infection timing
  t_infection = t_infection #time range of antigen presence (V = 1 when t in t_infection)
)

# simulation Time parameters 
Tmax   = 365*3                         # max time in days
points.per.day = 24              # output points per day

#initial conditions
init_cond = c("B" = 10^4/10^3, "P" = 0, "M_s" = 0)

# solve 
soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model2, parms = params)) 

# long form ode 
soln_long = soln %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
   labs(x = "days", y = bquote("log"[10] ~ "number"), title = "primary infection, inducible response, 1 PC population") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)
```

```{r plot-ODE-Model-2-secondary-infection, evaluate = F}
# set initial conditions
init_cond = soln %>% select(!(time)) %>% slice(nrow(soln)) %>% unlist()

# simulation Time parameters 
Tmax   = 3                         # max time in days
points.per.day = 24              # output points per day

# solve 
soln_secondary = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model2, parms = params)) 



# long form ode 
soln2_long = soln_secondary %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
  labs(x = "days", y = bquote("log"[10] ~ "number"), title = "secondary infection, inducible response, 1 PC population") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)
```

plotting $M_s$ together: 

```{r plot-together, evaluate = F}
plot = ggplot() + geom_line(data = soln2_long %>% filter(population == "M_s"), aes(x = time, y = log10(number), linetype = "inducible"), lwd = 1.5) + 
  geom_line(data = soln_notinducible2_long %>% filter(population == "M_s"), aes(x = time, y = log10(number), linetype = "not inducible"), lwd = 1.5) +
  scale_linetype_manual(values = c("inducible" = "solid", "not inducible" = "dashed")) +
  labs(x = "days", y = bquote("log"[10] ~ "number"), title = "secondary infection, antibody response") +
  theme_bw() #removes grey background and changes aesthetics

#print to viewer
print(plot)


```
