---
title: "Competition for Transport Constrains Respiratory IgA Longevity 3:"
subtitle: "Calculate: nonspecific IgA inhibits specific IgA translocation to site of infection"
author: Cora, Susi, Rustom, Hasan, Carl...
date: "`r Sys.Date()`"
format: html

---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
source(here("code", "functions", "functions.R"))
```

# Objective 

The goal of this document is to calculate the extent to which pre-exisiting, non-specific IgA antibodies interfere with he transport of newly generated, specific IgA across the mucosal epithelium (where it will be useful.)

# Outline

We can calculate the pre-existing, non-specific antibody titer in two ways: 

1) we can solve for the steady state analytically, given the equation from `\code\2_IgA_transport.qmd`. This saves time, space, and is simpler. Or, 
2) we can solve for the steady state by solving the ode describing the dynamics of IgA change in the mucosa over time, and extracting the average non-specific titer at steady state. This might be useful if we consider adding other compartments or more complex dynamics reflective of the biology of IgA that complicate the calculation of the steady-state concentration. 

# Simulation 


We want to vary the decay rate. $d_\mu$, and extent of competition, $K$; or vary the diffusion of non-specific IgA from serum. 
Lets calculate the ss for chaning d_mu: 

```{r simulated-parms}
#half lives for decay rates: 
t.half = c(1,2,5,10,25) #in years

#make list of parms
parms = list(
  Vmax = 10, #maximum rate of IgA transport
  K = 100, #concentration of IgA for which transport is half-max
  ## steady state concentration of non-specific IgA
  f = 10, #flux of IgA into mucosa from blood per day
  A0 = 100, #peak antibody titer for a mucosal response
  tau = 365, #time between respiratory infections in days
  gamma_m = A0/tau + f,  #constant flow of IgA into mucosal compartment due to local production and flux from the blood
  d_mu = -log(1/2)/(t.half*365), #decay rate of mucosal IgA in terms of half life, in days
  Vmax_0 = 10, # average max transport rate of IgA to lumen per day
  #newly generated IgA
  r = log(10^2)/14, # exponential growth of IgA on new infection
  rho = 14) #time of peak of response)


```

below we calculate the steady states for varying $d_\mu$ 

```{r ss_solution}
# define changing variable
d_mus = -log(1/2)/(c(1,2,5,10,25)*365)

#calculate from function
Mns_ss = sapply(d_mus, function(d) {Mns_SteadyState(A0 = parms$A0, tau = parms$tau, f = parms$f, Vmax_0 = parms$Vmax, d_mu = d, K = parms$K)})

print(Mns_ss)

```

and here I check (because that feels wrong):

```{r }
# define changing variable
d_mus = -log(1/2)/(c(1,2,5,10,25)*365)

## simulation parameters 
Tmax   = 100*365                  #20       # max time in years
points.per.day = 24                # output points per year
duration.of.steady.state = 5*365        # duration in days for averaging steady state for seasonal

## initial conditions 
init.cond = c(Mns = 1) # initial mucosal iga

#iterate through mucosal decay rates and solve
Mns_odes = lapply(d_mus, function(d_mu){
  #change d_mu parameter in the parms list
  parms$d_mu = d_mu
  
  #calculate dynamics
  soln = as.data.frame(ode(y = init.cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), Mns.odes.model, parms = parms))
  }
)


```

```{r plot-dynamics}
#plot dynamics
plot = ggplot() #create empty plot
for(soln in Mns_odes){ #iterate through model solutions
  plot = plot + 
    geom_line(data = soln, aes(y = Mns, x = time/365)) +
    geom_hline(yintercept = Mns_ss, linetype = "dashed") +
    geom_text(aes(x = c(t.half), y = Mns_ss, label = paste0("T[1/2]==", t.half, "~years")), parse = T, nudge_y = 100, nudge_x = 10) + #labels ss with corresponding thalf
    theme_bw()+          
    labs(y = "Mucosal IgA, Total", x = "Time, years")
}
print(plot)

#extract steady states
Mns_odes_SS = sapply(Mns_odes, function(soln){
  Mns = soln$Mns #extract vector of titer over time
  ss = mean(tail(Mns, duration.of.steady.state*points.per.day))
})

#compare ss from ode to ss from calculation
Mns_ss #from calculation
Mns_odes_SS #from ode numerical solution

```

Still not happy with the parms but at least the calculation is correct!

Below, we simulate a response over the course of month and observe the change in specific antibodies in the mucosa $M_S$ over time:

```{r specific-antibodies}
# response parameters
times = seq(0, 30, by = 1/24) #1 month, calculated by hour
# specific antibody titer over time
M_s = sapply(d_mus, function(d_mu) {
  sapply(times, function(t) {
    M_specific(A0 = parms$A0, r = parms$r, rho = parms$rho, d_mu = d_mu, t = t)
  })
}) %>% as.data.frame()

#add time column 
M_s = cbind(times, M_s) %>% as.data.frame()
colnames(M_s) = c("time", d_mus) #name columns for easy reference

M_s.longer = M_s %>%
  pivot_longer(cols = !time, 
               names_to = "decay",
               values_to = "Ms")

# plot dynamics
plot = ggplot() + geom_line(data = M_s.longer, aes(x = time, y = Ms, group = decay)) +
  theme_classic() +
  labs(x = "Time, days", y = "Specific Antibody in Mucosa, Ms")

print(plot)
```

```{r calculate-flux-over-time}
Phi.s_l = sapply(d_mus, function(d_mu){
  sapply(times, function(t) {
    #calculate ss M_ns
    M_ns = Mns_SteadyState(A0 = parms$A0, tau = parms$tau, f = parms$f, Vmax_0 = parms$Vmax, d_mu = d_mu, K = parms$K)
    #calculate M_s at that time
    M_s = M_specific(A0, r, rho, d_mu, t)
    
    # calculate flux of specific IgA
    return(specific_flux(Vmax = parms$Vmax, K = parms$K, M_ns = M_ns, M_s = M_s))
  })
}) %>%
  cbind(times, .) %>% # add time column
  as.data.frame() #format as data frame

# add column names
colnames(Phi.s_l) = c("time", t.half)

# pivot to long form 
Phi.s_long = Phi.s_l %>% 
  pivot_longer(cols = !time, 
               names_to = "t.half",
               values_to = "flux")

```

```{r plot-flux-over-time}
# plot flux for each decay rate over time
plot = ggplot() +
  geom_line(data = Phi.s_long, aes(x = time, y = flux, group = as.factor(as.numeric(t.half)), col = as.factor(as.numeric(t.half)))) +
  theme_bw() +
  labs(x = "days post infection", y = "flux of specific IgA", col = expression(paste("T"[1/2], ", in years")))

print(plot)
```
