---
title: "Evolutionary Constraints on the Longevity of IgA Antibody Titers"
subtitle: "Simulations Part 2: Rapid mucosal recall responses may be facilitated by inducible plasma cell populations"
format: html
---

# Objective

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
source(here("code", "functions", "functions.R")) #loads functions from functions file


```

# Objective

The goal of this document is to simulate the recall of antibody responses in the mucosa, including a population of inducible plasma cell which secrete antibody only when stimulated by antigen. We would like to determine the extent to which this mechanism can facilitate a rapid antibody response to a previously encountered antigen when the longevity of titers in the respiratory tract wane rapidly. 


# Visualizing dynamics

We will first compare the dynamics of antibody titer increase after a primary infection and a secondary infection with the same antigen. 

## Define parameters

```{r define-parms}
## Defining parameters for simulations 
  # B cell parms
    r_b = 2 #rate of proliferation of activated B cells, twice per day
    diff_frac = 0.003 #rate of differentiation of B cells into AS Plasma Cell, 5 % become ASCs
  # Plasma cell differentiation parameters
    quiessence_k = 0.003 #rate of differentiation of plasma cells into quiescent plasma cells
    induction_k = 1 #rate of differentiation of quiescent plasma cells to AS plasma cells
  # Antibody production and transport
    rho = 10^3 #rate of antibody secretion by plasma cells per day (scaled)
    Vmax = 10^3 #maximum rate of antibody transport to the mucosa
    K = 10^3 #Ab concentration at which transport rate is 1/2 Vmax
    c =  10^6 #rate of general antibody influx into mucosa per day - for calculating ss nonspecic M_ns
  # decay rates - Cells - defined in terms of half lives
   thalf_mu = 14 # in days
    d_mu = -log(1/2)/(thalf_mu) #AS plasma cell death rate 
   thalf_q = 5 # in years
    d_q = -log(1/2)/(thalf_q*365)  #quiescent cells death rate, per day
    d_B = d_q #B cell death rate
  # Decay rates - antibodies
   thalf_Ab = 10 # in days
    d_Ab = -log(1/2)/(thalf_Ab)  #decay rate of IgA in mucosa
   thalf_l = thalf_Ab # in days
    d_l = -log(1/2)/(thalf_l)  #decay rate of IgA in lumen 
```

Above, we define the basic parameters for our model system. 

Below, we oragnize these parameters into a list for calling in
`deSolve` and define the times for integration, as well as the intial conditions, or our ODE model. 

```{r primary-simulation-setup}
# params list for deSolve
params = list(
## Defining parameters for simulations 
  # B cell parms
    r_b = r_b, #rate of proliferation of activated B cells
    diff_frac = diff_frac, #rate of differentiation of B cells into AS Plasma Cell
  # Plasma cell differentiation parameters
    quiessence_k = quiessence_k, #rate of differentiation of plasma cells into quiescent plasma cells
    induction_k = induction_k, #rate of differentiation of quiescent plasma cells to AS plasma cells
  # Antibody production and transport
    rho = rho, #rate of antibody secretion by plasma cells
    Vmax = Vmax, #maximum rate of antibody transport to the mucosa
    K = K, #Ab concentration at which transport rate is 1/2 Vmax
    c = c, #rate of general antibody influx into mucosa - for calculating ss nonspecic M_ns
  # decay rates - Cells
    d_mu = d_mu, #AS plasma cell death rate 
    d_q = d_q, #quiescent cells death rate
    d_B = d_B, #B cell death rate
  # Decay rates - antibodies
    d_Ab = d_Ab, #decay rate of IgA in mucosa
    d_l = d_l, #decay rate of IgA in lumen 
    t_infection = c(0,14)
  )
## For simulation 
  # for time vector
  Tmax = 120 # max time (days) to simulate over
  points_per_day = 24 # number of points per day to simulate over

# initial conditions
  init_cond = c(B = 2, # Initial pathogen-specific B cells
                P_s = 0, # no pathogen-specific antibody secreting plasma cells initially
                P_q = 0, # no pathogen-specific quiescent  plasma cells
                M_s = 0, # no specific antibodies in mucosa 
                L_s = 0) # no specific antibodies in lumen 
# M_ns is calculated in ODE model function

```

## Run simulation

```{r }
soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), secondary_response.model, parms = params))

soln.long = soln %>% pivot_longer(cols = !time, 
                                  names_to = "population", 
                                  values_to = "number")

plot1 = ggplot() + 
  geom_line(data = soln.long %>% filter(population %in% c("B", "P_s", "P_q")), aes(x = time, y = log10(number), group = population, col = population))

plot2 = ggplot() + 
  geom_line(data = soln.long %>% filter(population %in% c("M_s", "L_s")), aes(x = time, y = log10(number), group = population, col = population))

grid.arrange(plot1,plot2)
```

