---
title: "1_primary_responses_Solution"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
source(here("code", "functions", "functions.R"))

```

# Objective 

In this document, we evaluate how long-lived IgA antibody titers in the respiratory mucosa interfere with the transport of specific IgA antibodies to the site of infection. 

# How to measure efficiency 

We might be interested in the effect of competition on a few different quantities: 

  1) The concentration of specific antibody that makes it to the lumen, $L_s$, over time, whose dynamics are described by: 

$$
\frac{dL_s}{dt} = k_2R_s - d_L L_s
$$
  
  2) The relative proportion of transport proteins carrying specific antibody out of the total number of transport proteins carrying antibody, given by:

$$
f_s = \frac{R_s}{R_s + R_{ns}}
$$

  3) The efficiency of transport of specific antibody relative to the amount of mucosal antibody, given by: 

$$
\epsilon_s = \frac{k_2R_s}{M_s}
$$

These might all be plotted as a function of time (x-axis) with various lines describing the dynamics given different $d_\mu$. 

However, we might also want to plot a heatmap of the effect of saturation on the response given changes in $d_\mu$ and strength of competition (either $K_M$ or $V_{max}$ changing). For these heatmaps, we will need a single value to report. My suggestion is either to report one of these values at the peak of the specific response ($t_{peak}$), or to plot the integral of one of these values up until the peak of the response. 

I will make that decision after seeing the results of the dynamics for each measurement! 

# Parameters 

Below, we define the parameters required for the model `IgA_competition.model` in `code\functions\function.r`. This model is described in `code\simulations\1_primary_responses_Outline.qmd`. Importantly, we define a vector of $d_\mu$ to identify the effect of larger and larger concentration of total IgA on inhibitin the transport of newly generated IgA to a novel pathogen. 

## Antibody concentrations 

First, we define how we vary the concentration of nonspecific antibody $M_{ns}$ with respect to the dynamics of specific antibody $M_s$ increase and decay during infection.

Importantly, the steady state concentration of $M_{ns}$ is not simply the steady state of $\frac{dM_{ns}}{dt} = c - d_\mu$ which gives $\frac{c`}{d_\mu}$, because antibody is constantly being transported into the lumen. Instead, the equilibrium concentration of $M_{ns}$ should be derived as the positive solution to the steady state of the following system of ODEs:

$$
\frac{dM_{ns}}{dt} = c - k_1 (R_T - R_{ns}) M_{ns} - d_\mu M_{ns}
$$

and

$$
\frac{dR_{ns}}{dt} = k_1(R_T - R_{ns})M_{ns} - k_2R_{ns}
$$

solving for the steady state gives: 

$$
M_{ns}^{*} = \frac{-\left(d_{\mu}k_{2}+k_{1}k_{2}R_{T}-ck_{1}\right)+\sqrt{\left(d_{\mu}k_{2}+k_{1}k_{2}R_{T}-ck_{1}\right)^{2}+4d_{\mu}k_{1}ck_{2}}}{2d_{\mu}k_{1}}
$$


Below, we plot the dependence of $M_{ns}^{*}$ on the decay rate ($d_\{mu}$). 

```{r define-parameters-transport}
# transport
Vmax = 10^3 # order of magnitude of peak response # we might vary this later
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
k1 = k2/10^2 #k2/min(Mns_vec) #rate of antibody binding such that KM is order of magnitude of min of the Mns
KM = k2/k1 #checking KM

# dynamics in the lumen 
t_half_L = 7 #half life of antibodies in the lumen, in days
d_L = log(2)/t_half_L #decay rate in the lumen

# steady state nonspecific antibodies
c = 1*10^5 #rate of influx of nonspecific IgA, Ab/time - comes from pb/dA - in mg/dl per day
D = 0 #diffusion out
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years
d_mus = log(2)/t_halfs #half lives in days
names(d_mus) = t_halfs

```

```{r steady-state-Mns}
#define steady state concentration of mns
ss_Mns = function(d_mu) {(-(d_mu*k2 + k1*k2*RT - c*k1) + sqrt((d_mu*k2 + k1*k2*RT - c*k1)^2) + 4*d_mu*k1*c*k2)/2*d_mu*k1}

# calculate ss concentrations 
sapply(d_mus, function(d_mu){ss_Mns(d_mu = d_mu)})

#plot
plot = ggplot() + geom_hline(yintercept = sapply(d_mus, function(d_mu){ss_Mns(d_mu = d_mu)}))

plot
```



```{r define-parameters-antibody-production}
### parameters for competition model 
# steady state nonspecific antibodies
c = 1*10^5 #rate of influx of nonspecific IgA, Ab/time - comes from pb/dA - in mg/dl per day
D = 0 #diffusion out
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years
d_mus = log(2)/t_halfs #half lives in days
names(d_mus) = t_halfs
Mns_vec = c/(D+d_mus) #vector of the ss nonspecific mucosal antibodies

# generation of new response
Ap = unname(c/(D + d_mus["90"])) #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = Ap/10^3 #1 1000th of the peak is where we start
tp = 5 #time of peak titer -- days

# will be calculated when iterating through d_mu
#r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
#alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

#plot of change in r for specific cell production
for(d_mu in d_mus){
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate
  if(d_mu == d_mus[1]){
  curve(r0 - alpha*x, from = 0, to = tp, xlab = "time, days", ylab = "growth rate of Ms", main = "changing growth rate of IgA population until peak")}
  else {curve(r0 - alpha*x, from = 0, to = tp, add = T)}
}
```

Below, we plot how the dynamics of specific antibody generation and decay compare to the steady-state concentrations of nonspecific antibody over the course of infection. 

```{r plot-Ms-v-Mns}
### plot Ms vs Mns over time
Mab_titers = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = c/(D+d_mu) # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + D + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu - D)/tp # rate of decay of exponential growth rate

  ### deSolve for specific ab over time
  
  # parms list
  params = list( # parameter list
  r0 = r0, 
  A0 = A0,
  Ap = Ap,
  alpha = alpha,
  tp = tp,
  d_mu = d_mu, 
  D = D
  )

  # initial conditions
  Ms_init = A0 # starting antibody titer
  init_cond = c("Ms" = Ms_init)

  # numerical solver parameters
  Tmax = 30 # total days
  points_per_day = 24 # every hour

  # solve 
  soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), primary_response.model, parms = params)) %>% cbind(.,Mns, t_half)
}) %>% do.call(rbind, .) #concatenates all dfs from lapply list into a single df rowwise

### Plot
plot = ggplot() + geom_line(data = Mab_titers, aes(x = time, y = log10(Ms), group = factor(t_half), col = factor(t_half), linetype = "Ms")) +
  geom_line(data = Mab_titers, aes(x = time, y = log10(Mns), group = factor(t_half), col = factor(t_half), linetype = "Mns")) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("Ms" = "solid", "Mns" = "dashed")) +
  labs(x = "time, days", y = "log10[Ab], mg/dL", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "[Ms] and [Mns] over the course of infection") +
  theme_bw()
  

#print to viewer
print(plot)

```

## Antibody transport 

Now, we define our parameters for the transport of antibodies across mucosal epithelium, as well as the decay of antibody in the lumen: 

```{r define-parameters-transport}
# transport
Vmax = Ap # order of magnitude of peak response # we might vary this later
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
RT = 1#Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
k1 = 1 #k2/min(Mns_vec) #rate of antibody binding such that KM is order of magnitude of min of the Mns
KM = k2/k1 #checking KM

# dynamics in the lumen 
t_half_L = 7 #half life of antibodies in the lumen, in days
d_L = log(2)/t_half_L #decay rate in the lumen

```

and plot how the flux of $M_s$ into the lumen depends on the steady state concentration of $M_{ns}$ and the concentration of $M_s$

```{r plot-Ms-flux-curve-by-Mns}
### plot flux vs [Ms] given each Mns 
flux_df = lapply(t_halfs, function(t_half){
  # set d_mu and Mns 
  d_mu = log(2)/t_half
  Mns = c/(d_mu + D)
  
  #set concentrations of Ms
  concentrations = 10^seq(-3, log10(max(Mns_vec)), by = 0.1)
  
  #calculate flux
  flux = sapply(concentrations, function(Ms){
    Vmax*Ms/((k2/k1) + Mns + Ms)
  }) %>% cbind("Ms" = concentrations, "flux" = ., "t_half" = t_half) %>% as.data.frame() 
}) %>% do.call(rbind, .) %>% as.data.frame() 

## plot
plot = ggplot() + geom_line(data = flux_df, aes(x = Ms, y = flux, group = factor(t_half), col = factor(t_half))) +
  geom_hline(yintercept = Vmax, linetype = "dashed") +
  geom_vline(xintercept = k2/k1, linetype = "dashed") +
#  geom_vline(xintercept = k2/k1 + Mns_vec, linetype = "dotted") +
  geom_vline(xintercept = Ap, color = "red") +
  scale_color_manual(values = my_colors) +
  labs(x = "[Ms]", y = "flux [Ms]", color = "Mucosal Ab Half Life, days", title = "Competitive inhibition of specific antibody transport by nonspecific antibodies") + 
  theme_bw()

# print to viewer
print(plot)
```

The horizontal dashed line shows the greatest possible transport rate, $V_{max}$, the vertical dashed line shows the 50% saturation concnetration given $[M_s]$ only, and I have added a red vertical line to show the maximum concentration of $M_s$ over an infection.

RETURN TO THIS TO ADD JUSTIFICATION FOR PARMS 

# Dynamics as a function of mucosal antibody half-life

We would like to show how high concentrations of nonspecific antibody in the mucosa interfere with the transport of specific antibody across the mucosa and into the lumen, the site of infection. 

## Concentration of specific antibody, $L_s$ in the lumen over time

Here, we numerically solve for the concentration of specific antibody in the lumen over time: 


```{r ode-setup-IgA-competition}
### parameters for competition model 
## define half-lifes
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years

### define initial conditions and parms
# initial conditions
  Mns_init = 0
  Rns_init = 0
  Lns_init = 0
  init_cond = c("Mns" = Mns_init, "Rns" = Rns_init, "Lns" = Lns_init)

  # numerical solver parameters
  Tmax = 100000 # total days
  points_per_day = 1 # every hour


### deSolve soltions
ss_solns = lapply(t_halfs, function(t_half){ # half lives in days
  ### set parms 
    d_mu = log(2)/t_half #decay rate per day
    
  # As list
  params = list( #setting A
    # nonspecific ss
    d_mu = d_mu, # plasma cell decay rate per day
    c = c, # total antibody production rate per day
    D = D,
    # transport
    k2 = k2, # trancytosis rate
    RT = RT, # total number of receptors
    k1 = k1, # ab binding rate
    # lumen
    d_L = d_L #decay rate in the lumen
  )
  
  # ode solve for steady state without specific response (to account for transport of nonspecific to lumen)
  soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), equilibrium_ns.model, parms = params)) %>% cbind(., t_half)
}) %>% do.call(rbind, .) %>% pivot_longer(cols = !c(time, t_half),
                                          names_to = "population",
                                          values_to = "concentration")


plot = ggplot() +
  geom_line(data = ss_solns %>% filter(population %in% c("Mns"), time >= Tmax-30), aes(x = time, y = log10(concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = "Mns")) +
  geom_line(data = Mab_titers, aes(x = time + Tmax-30, y = log10(Ms), group = factor(t_half), col = factor(t_half), linetype = "Ms")) +
  scale_color_manual(values = my_colors) +
  theme_bw()

# print to viewer
print(plot)

```

This is maybe a better approximation for the regime we are in...



```{r ode-setup-IgA-competition}
### parameters for competition model 
## define half-lifes
t_halfs = c(14,30,90,400, 4000, 20000) #half lives of mucosal ab in days - 2 weeks, 1 month, 3 months, 1 year, 10 years, 50 years

### define initial conditions and parms
# initial conditions
  Ms_init = 0 # starting antibody titer
  Mns_init = 0
  Rs_init = 0
  Rns_init = 0
  Ls_init = 0
  Lns_init = 0

### numerical simulation parms
Tmax_ss = 100000
points_per_day_ss = 1

Tmax = 100
points_per_day = 24 

### deSolve soltions
competition_solns = lapply(t_halfs, function(t_half){ # half lives in days

  ############ solve for steady state concentrations ###################
  init_cond = c("Ms" = 0, "Mns" = Mns_init, "Rs" = 0, "Rns" = 0, "Ls" = 0, "Lns" = Lns_init)
  ### set parms 
    d_mu = log(2)/t_half #decay rate per day
    
  # As list
  params = list( #setting A
    # nonspecific ss
    d_mu = d_mu, # plasma cell decay rate per day
    c = c, # total antibody production rate per day
    D = D, 
    #generation of new responses
    Ap = Ap, # peak antibody titer
    A0 = A0, # intial antibody titer
    tp = tp, # time of peak response, days
    r0 = (2*log(Ap/A0)/tp) + d_mu, #initial/greatest exponential expansion rate
    alpha = (r0 - d_mu)/tp, # rate of decay of exponential growth rate
    # transport
    k2 = k2, # trancytosis rate
    RT = RT, # total number of receptors
    k1 = k1, # ab binding rate
    # lumen
    d_L = d_L #decay rate in the lumen
  )
  
  # ode solve for steady state without specific response (to account for transport of nonspecific to lumen)
  init_cond = as.data.frame(ode(y = init_cond, times = seq(0, Tmax_ss, length.out = Tmax_ss*points_per_day_ss), IgA_competition.model, parms = params)) %>% tail(.,1) %>% select(!time) %>% unlist(.)
  
  ############## course of infection ######################
  init_cond["Ms"] = A0 # set specific antibody to increase
  
  soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points_per_day), IgA_competition.model, parms = params)) %>% cbind(., t_half) 
  
  return(soln)
}) %>% do.call(rbind, .) %>% pivot_longer(cols = !c(time, t_half),
                                          names_to = "population",
                                          values_to = "concentration")


plot = ggplot() +
  geom_line(data = competition_solns %>% filter(population %in% c("Lns", "Ls")), aes(x = time, y = log10(concentration), group = interaction(factor(t_half), population), color = factor(t_half), linetype = population)) +
  scale_y_continuous(limits = c(-5,5)) +
  scale_color_manual(values = my_colors) +
  theme_bw()

# print to viewer
print(plot)

```

