---
title: "2_secondary_responses_simplified"
format: html
editor: visual
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
#source(here("code", "functions", "functions.R"))

#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective

The goal of this document is to explore a biological mechanism of IgA recall responses in the respiratory tract - namely that, rather than constitutively secreting antibody, localized IgA plasma cells are inducible - through ODE models. 

# Outline 

I plan to formulate 2 models that both result in inducibility: 

1) Once plasma cells are generated, they differentiate into "quiescent" plasma cells at some rate. On recall responses, they are recalled at some rate. 

2) Once plasma cells are generated, they only secrete antibody in the presence of antigen. When antigen disappears, secretion stops. 

I think the first will be more gradated/biological for a model with bianry "infection/non-infection". But, if we allow the rate of recruitment (1) or the rate of secretion (2) to be dependent on the density of antigen, then I'm not entirely sure how different the two models will be!

# Model 1 - Plasma cell differentiation

Below, we define an ode model in which there are 3 populations of cells: B cells, secretory plasma cells, and quiescent plasma cells. We include a fourth variable - "Virus" - which is either 0 (no infection present) or 1 (infection present). 

When $V = 0$, each cell population decays at some rate, and secretory plasma cells differentiate into quiescent plasma cells. 

When $V = 1$, B cells proliferate and differentiate into secretory plasma cells. Secretory plasma cells differentiate into quiescent plasma cells at some rate. Quiescent plasma cells are recruited to differentiate into secretory plasma cells at some rate. Cell populations continue to decay according to their given death rate. 

```{r define-ODE-Model-1}
# Antibody concentrations in the respiratory mucosa #
secondary_response.model1 <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        B = y[1]         # B cells 
        P_s = y[2]       # Antibody secreting cells (specific)
        P_q = y[3]       # quiescent antibody secreting cells (specific)
        M_s = y[4]       # specific Ab in mucosa
        
        # make empty variables for the derivatives
        dB = 0      # change in specific Ab in the lumen
        dP_s = 0    # change in specific ASCs
        dP_q = 0    # change in quiescent ASCs
        dM_s = 0    # change in specific antibodies in mucosa
        
        #####################################
        # define V function 
        V.function = function(t_infection, t) {
          if(t>=min(t_infection) & t <= max(t_infection)){
            V = 1}
          else(V = 0)
        return(V)}
        # set V
        V = V.function(t_infection, t)
        #####################################
        ## calculate the derivatives
        # cell populations
        dB = r*(1-f)*V*B - d_B*B #proliferation only when virus present, some fraction f differentiates to P_s
        dP_s = r*f*V*B + s*V*P_q - d_s*P_s - q*P_s #differentiation from B cells or P_q only when virus present
        dP_q = q*P_s - d_q*P_q - s*V*P_q #P_s always differentiates to P_q with some rate q.
        
        # Specific antibodies        
        dM_s = a*P_s - d_a*M_s #P_s secretes antibody
        
        # output the derivatives
        dy=c(dB, dP_s, dP_q, dM_s) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} # we will need to think about how to set M_ns if we want to track specific IgA in the lumen.

```

Instead of simulating a primary infection followed by a secondary infection - because we really don't know what the "rules" are for feedback between antibody and control of infection/antigen load, and subsequently generated responses - we will instead initiate a secondary response with some ratio of B cells to quiescenet cells/ B cells to inducible plasma cells. This ratio will be set by fraction of B cell divisions that give rise to plasma cells; we assume that the quiescent cells decay with the same rate as the memory B cells, and so their ratio should stay the same over time. 

### Model 1 - quiescent cell differentiation

```{r set-parms-ODE-Model-1}
# death rates
d_B = log(10)/(365) # B cell death rate, x days half life
d_s = 0 # secretory cell death rate #don't consider for now
d_q = log(10)/(365) # quiescent cell death rate long lived like bm
  
# proliferation and differentiation
r = 3 # b cell proliferation rate - doubling time is 1/2 day
f = 1/3 # fraction of division that produce secretory cells
q = log(2)/14 # rate of secretory cells to quiescence #should be dmu 
s = 3 # rate of quiescent cell recruitment to secretion #should be very fast, like a percentage per hour

# antibody dynamics
a = 10 # rate of IgA secretion by secretory cells # not dealing with density but total number now
d_a = log(2)/3 # IgA molecular decay rate, 3 days half life

# infection timing
t_infection = c(0,14) #time range of antigen presence (V = 1 when t in t_infection)
```


```{r plot-ODE-Model-1-secondary-infection}
# parms list 
# parameters list for deSolve
params = list(
  d_B = d_B, # B cell death rate
  d_s = d_s, # secretory cell death rate
  d_q = d_q, # quiescent cell death rate
  d_a = d_a, # IgA molecular decay rate
    
  # proliferation and differentiation
  r = r, # b cell proliferation rate
  f = f, # fraction of division that produce secretory cells
  q = q, # rate of secretory cells to quiescence
  s = s, # rate of quiescent cell recruitment to secretion
  
  # antibody dynamics
  a = a, # rate of IgA secretion by secretory cells
  d_a = d_a, # IgA molecular decay rate in mucosa
  
  #infection timing
  t_infection = t_infection #time range of antigen presence (V = 1 when t in t_infection)
)

# set initial conditions
#initial conditions
init_cond = c("B" = 3, "P_s" = 0, "P_q" = 1, "M_s" = 0)

# simulation Time parameters 
Tmax   = 3                       # max time in days
points.per.day = 24              # output points per day

# solve 
soln_secondary = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model1, parms = params)) 



# long form ode 
soln2_long = soln_secondary %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
  coord_cartesian(ylim = c(0,4)) +
  labs(x = "days",, y = bquote("log"[10] ~ "number"), title = "secondary infection, inducible response") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)
```

compare to if $d_\mu$ was actual death rate

```{r ODE-model-no-inducible}
# Antibody concentrations in the respiratory mucosa #
secondary_response.model_notinducible <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        B = y[1]         # B cells 
        P_s = y[2]       # Antibody secreting cells (specific)
        M_s = y[3]       # specific Ab in mucosa
        
        # make empty variables for the derivatives
        dB = 0      # change in specific Ab in the lumen
        dP_s = 0    # change in specific ASCs
        dM_s = 0    # change in specific antibodies in mucosa
        
        #####################################
        # define V function 
        V.function = function(t_infection, t) {
          if(t>=min(t_infection) & t <= max(t_infection)){
            V = 1}
          else(V = 0)
        return(V)}
        # set V
        V = V.function(t_infection, t)
        #####################################
        ## calculate the derivatives
        # cell populations
        dB = r*(1-f)*V*B - d_B*B #proliferation only when virus present, some fraction f differentiates to P_s
        dP_s = r*f*V*B - d_s*P_s #differentiation from B cells or P_q only when virus present
        
        # Specific antibodies        
        dM_s = a*P_s - d_a*M_s #P_s secretes antibody
        
        # output the derivatives
        dy=c(dB, dP_s, dM_s) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} # we will need to think about how to set M_ns if we want to track specific IgA in the lumen.
```

```{r set-parms-ODE-Model-1-noinducible}
# death rates
d_B = log(2)/(365) # B cell death rate, 2 years half life
d_s = q # rate of secretory cells to death #should equal d_mu
 
# proliferation and differentiation
r = 3 # b cell proliferation rate - doubling time is 1 day
f = 1/3 # fraction of division that produce secretory cells

# antibody dynamics
a = 10 # rate of IgA secretion by secretory cells # not dealing with density but total number now
d_a = log(2)/3 # IgA molecular decay rate, 3 days half life

# infection timing
t_infection = c(0,14) #time range of antigen presence (V = 1 when t in t_infection)
```

```{r solve-ODE-Model-1-first-infection-noinducible}
# parameters list for deSolve
params = list(
  d_B = d_B, # B cell death rate
  d_s = d_s, # secretory cell death rate

    
  # proliferation and differentiation
  r = r, # b cell proliferation rate
  f = f, # fraction of division that produce secretory cells

  # antibody dynamics
  a = a, # rate of IgA secretion by secretory cells
  d_a = d_a, # IgA molecular decay rate in mucosa
  
  #infection timing
  t_infection = t_infection #time range of antigen presence (V = 1 when t in t_infection)
)
```

```{r plot-ODE-Model-1-secondary-infection-noinducible}
# set initial conditions
init_cond = soln_notinducible %>% select(!(time)) %>% slice(nrow(soln_notinducible)) %>% unlist()

# simulation Time parameters 
Tmax   = 30                       # max time in days
points.per.day = 24              # output points per day

# solve 
soln_notinducible_secondary = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model_notinducible, parms = params)) 



# long form ode 
soln_notinducible2_long = soln_notinducible_secondary %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln_notinducible2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln_notinducible2_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
   labs(x = "days", y = bquote("log"[10] ~ "number"), title = "secondary infection, not inducible response") +
   coord_cartesian(ylim = c(0,3)) +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)
```
