---
title: "Constraints on the longevity of IgA: Simulations of specific IgA transport in the context of pre-existing, nonspecific immunity"
subtitle: "Paper Figures"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
MM_function = function(M_s, M_ns) {(Vmax*M_s)/(KM+M_ns+M_s)}


#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective 

In part 1 this document, we evaluate how long-lived IgA antibody titers in the respiratory mucosa interfere with the transport of specific IgA antibodies to the site of infection. We assume that the concentration of $M_{ns}$ changes negligibly over the course of infection, and assume that the steady state concentration of $M_{ns}$ is proportional to $\frac{c}{d_\mu}$. We use a Michaelis-Mentin approximation to estimate the flux of specific IgA across the mucosal membrane. 


# Outline 

I will review the kinetics inferred by the Michaelis-Mentin approximation before presenting the results of competitive inhibition by non-specific immunity during a specific response. 

The results in this document are organized as follows: 

1) First, we will explore the effect of changing $d_\mu$ on the total flux of specific IgA, $k_2R_S$ (or $\phi_S$) over time. We will demonstrate how much specific IgA must be generated during a response to overcome the effect of competition from pre-existing titers. 



# Competition for transport: Michaelis-Mentin Kinetics

Below, we define our parameters: 

```{r define-parameters-MM}
# generation of new response
Ap = 100 #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = 1 # initial specific ab concentration
tp = 30 #time of peak titer -- days

# steady state nonspecific antibodies
t_halfs = c(30,90,400,4000) #half lives of mucosal ab in days - 1 month, 3 months, 1 year, 10 years :) 
d_mus = log(2)/t_halfs #half lives in days
  names(d_mus) = t_halfs
c = Ap*(log(2)/min(t_halfs))# calculate c such that the minimum M_ns = Ap = 100
Mns_vec = c/d_mus #vector of approximate ss solutions
  names(Mns_vec) = t_halfs


# transport
Vmax = 10^3 # order of magnitude of peak response # we might vary this later
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
KM = Ap/3 #choosing KM such that at peak in the absence of Mns, Ms would achieve 75% of max flux at peak
k1 = k2/KM #k2/min(Mns_vec) #rate of antibody binding such that KM is on the order of 10

# dynamics in the lumen 
t_half_L = 3 #half life of antibodies in the lumen, in days
d_L = log(2)/t_half_L #decay rate in the lumen


# Total time to plot responses
Tmax = 90

# will be calculated when iterating through d_mu
#r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
#alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate


```


Below, we plot how the flux of specific antibody for a given concentration of $M_s$ in the mucosa changes as a function of the half-life of non-specific antibodies. In essence, the effective amount of $M_s$ required to reach half-maximal transport will increase by the concnetration of non-specific antibody conferred by the increasing half-life. 

```{r plot-MM-kinetics}
#set concentrations of Ms for x axis
concentrations = 10^seq(-1, 4, by = 0.1) #range of concentrations M_s such that we span much less than and much greater than Ap

### plot flux vs [Ms] given each Mns 
flux_df = lapply(t_halfs, function(t_half){
  # set d_mu and Mns 
  d_mu = log(2)/t_half
  Mns = c/d_mu
  
  #calculate flux
  flux = sapply(concentrations, function(Ms){
   
   return(MM_function(M_s = Ms, M_ns = Mns))
  }) %>% cbind("Ms" = concentrations, "flux" = ., "t_half" = t_half) %>% as.data.frame() 
}) %>% do.call(rbind, .) %>% as.data.frame() 

## plot
plot = ggplot() + geom_line(data = flux_df, aes(x = (Ms), y = flux, group = factor(t_half), col = factor(t_half))) +
  geom_hline(yintercept = Vmax, linetype = "dashed") +
  geom_vline(xintercept = (KM), color = "red") +
  geom_vline(xintercept = (Ap), linetype= "dashed") +
geom_line(
  aes(x = concentrations, y = sapply(concentrations, function(Ms){Vmax*Ms/(KM+Ms)})),
  inherit.aes = FALSE,
  color = "red"
) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = my_colors) +
  labs(x = "[Ms]", y = "flux [Ms]", color = "Mucosal Ab Half Life, days", title = "Competitive inhibition of specific antibody transport by nonspecific antibodies") + 
  theme_bw()

# print to viewer
print(plot)
```

The horizontal dashed line shows the greatest possible transport rate, $V_{max}$; the red curve shows the MM kinetics given $[M_s]$ only; the vertical red line shows the 50% saturation concentration given $[M_s]$ only, $K_M$; and I have added a vertical dashed line to show the peak concentration of $M_s$ generated over an infection.

Notice that I have chosen paramerters such that, in the absence of competition, the maximum flux is reached by the the maximum concentration of $M_s$ generated during a response. As nonspecific antibody is added, however, the flux achieved at the maximum specific antibody concentration drops. This is the core of the competitive effect. 

Below, I show the generation and decay of $M_s$ over the course of infection with respect to the steady-state concentration of $M_{ns}$ for increasing half-life of mucosal antibodies. 

```{r plot-Ms-and-Mns}
### plot Ms vs Mns over time
Mab_titers = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = c/d_mu # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  # set time vector parms
  Tmax = Tmax # total days
  points_per_day = 24 # every hour
  time = seq(0, Tmax, length.out = Tmax*points_per_day) #generate time vector

  ### apply to calculate
  soln = data.frame("time" = time, "Ms" = sapply(time, function(t){
    if(t<=tp){A0*exp((r0 - d_mu)*t - (alpha/2)*t^2)}
    else {Ap*exp(-d_mu*(t-tp))}
  }))
  

  # solve 
  soln = soln %>% cbind(.,Mns, t_half)
}) %>% do.call(rbind, .) #concatenates all dfs from lapply list into a single df rowwise

### Plot
plot = ggplot() + geom_line(data = Mab_titers, aes(x = time, y = log10(Ms), group = factor(t_half), col = factor(t_half), linetype = "Ms"), lwd = 1.5) +
  geom_line(data = Mab_titers, aes(x = time, y = log10(Mns), group = factor(t_half), col = factor(t_half), linetype = "Mns"), lwd = 1.5) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("Ms" = "solid", "Mns" = "dashed")) +
  labs(x = "time, days", y = "log10[Ab], mg/dL", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "[Ms] and [Mns] over the course of infection") +
  theme_bw()
  

#print to viewer
print(plot)

```

I have chosen the peak of the response such that the peak titer matches the steady state of the nonspecific antibody for an intermediate half-life. 

# Results 1: Flux over time as a function of $d_\mu$ or $\frac{k_2}{k_1}$

The flux of specific IgA over time will depend on competition with pre-existing immunity. This competition changes according to i) the amount of non-specific pre-exisitng immunity, which is a function of the rate of decay of responses, and ii) the magnitude of the peak specific antibody titer with respect to the saturation of transport, which we can change as a funtion of $K_M$, the half saturation point, which equals $\frac{k_2}{k_1}$. 

## Flux over time as a function of $d_\mu$

Below, we calculate the flux of specific IgA over time as a function of the decay rate of previously generated IgA: 

```{r parms-flux-decay-rate}
# set parms 
#transport 
Vmax = Vmax
k2 = k2
RT = RT
k1 = k1
KM = KM

#lumen
t_half_L = t_half_L
d_L = d_L

#steady state Ab
c = c
t_halfs = t_halfs
d_mus = d_mus
Mns_vec = Mns_vec
Ap = Ap

# generating new response
Ap = Ap
A0 = A0
tp = tp

# Total time 
Tmax = Tmax
```

```{r calculate-flux-decay-rate}
MM_soln = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = c/d_mu # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  # set time vector parms
  Tmax = Tmax # total days
  points_per_day = 24 # every hour
  time = seq(0, Tmax, length.out = Tmax*points_per_day) #generate time vector

  ### apply to calculate Ms titers over time
  Ms_soln = data.frame("time" = time, "t_half" = t_half, "Ms" = sapply(time, function(t){
    if(t<=tp){A0*exp((r0 - d_mu)*t - (alpha/2)*t^2)}
    else {Ap*exp(-d_mu*(t-tp))}
  }), "Mns" = Mns)
  
  ### apply to calculate Ls flux over time
  Ms_soln$phi_s = sapply(Ms_soln$Ms, function(Ms){MM_function(M_s = Ms, M_ns = Mns)})
  Ms_soln$phi_ns = sapply(Ms_soln$Ms, function(Ms){(Vmax*Mns)/(KM+Ms+Mns)})
  
  return(Ms_soln)
}) %>% do.call(rbind, .)

```

```{r plot-flux-decay-rate}
#ggplot
plot = ggplot() + geom_line(data = MM_soln, aes(x = time, y = (phi_s), group = factor(t_half), col = factor(t_half), linetype = "phi_s"), lwd = 1.5) +
  geom_line(data = MM_soln, aes(x = time, y = (phi_ns), group = factor(t_half), col = factor(t_half), linetype = "phi_ns"), lwd = 1.5) + 
  scale_color_manual(values = my_colors) +
  #coord_cartesian(ylim = c(0, 3), xlim = c(0, 15)) +
  scale_linetype_manual(values = c("phi_s" = "solid", "phi_ns" = "dashed"), labels = c("phi_s" = expression(phi[s]), "phi_ns" = expression( phi[ns]))) +
  labs(x = "time, days", y = bquote(phi[Ab] ~ ", day"^-1), col = bquote("T"[1/2] ~ ", days"), linetype = "Specificity") +
  theme_bw()+
  theme(axis.text = element_text(size = 25),
        axis.title = element_text(size = 25),
        legend.title = element_text(size = 25),
        legend.text = element_text(size = 25, hjust = 0),
        legend.key.width = unit(0.06, "npc")) 

#print to system viewer
print(plot)

#save
file = here("results", "competition", "fig-flux_over_time.png")
ggsave(plot, file = file)
```

yay

Now, we would like to plot how the peak flux (or the total amount of antibody transported) changes as a function of $t_{\frac{1}{2}}$ and the magnitude of the response in the mucosa. This way, we can see how the magnitude of the response required increases with $t_{1}{2}$

```{r parms-peak-decay-flux-c-matters}
# generation of new response
Aps = 10^seq(1, 3, by = 0.01) #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = 1 # initial specific ab concentration
tp = 30 #time of peak titer -- days

# steady state nonspecific antibodies
t_halfs = 3*10^seq(1,3.1, by = 0.01) #half lives of mucosal ab in days - 1 month, 3 months, 1 year, 10 years :) 

 
heatmap_matrix = sapply(Aps, function(Ap){ #columns
  sapply(t_halfs, function(t_half){ #rows
    #set other parms
    d_mu = log(2)/t_half #half lives in days
    c = Ap*(log(2)/min(t_halfs))# calculate c such that the minimum M_ns = Ap = 100
    Mns = c/d_mu #vector of approximate ss solutions
    
    # transport
    Vmax = 10^3 # order of magnitude of peak response # we might vary this later
    k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
    RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
    KM = Ap/3 #choosing KM such that at peak in the absence of Mns, Ms would achieve 75% of max flux at peak
    k1 = k2/KM #k2/min(Mns_vec) #rate of antibody binding such that KM is on the order of 10
    
    # dynamics in the lumen 
    t_half_L = 3 #half life of antibodies in the lumen, in days
    d_L = log(2)/t_half_L #decay rate in the lumen

    # Total time to plot responses
    Tmax = 90
    
    #calculate peak flux
    phi_s = MM_function(M_s = Ap, M_ns = Mns)
  })
})

### plot
### plot on log10 scale
x_ticks = c(30, 100, 300, 1000, 3000)   # half-life ticks (days)
y_ticks = c(10, 30, 100, 300, 1000)     # peak titer ticks
zlog <- log10(heatmap_matrix)
lev  <- pretty(range(zlog, finite = TRUE), 4)   # 20 = filled.contour's default nlevels

png(here("test2.png"), width = 2000, height = 1200, res = 300, type = "cairo", antialias = "none")
filled.contour(
  x = log10(t_halfs),
  y = log10(Aps),
  z = log10(heatmap_matrix),
  xlab = expression(paste("Plasma cell half-life, ", days)),
  ylab = "Peak specific antibody response in mucosa",
  main = "Peak Flux of Specific Antibody into Lumen",
  levels = lev,
  plot.axes = {
    axis(1, at = log10(x_ticks), labels = x_ticks)
    axis(2, at = log10(y_ticks), labels = y_ticks)
    contour(
      x = log10(t_halfs),
      y = log10(Aps),
      z = log10(heatmap_matrix),
      levels = lev,   # the flux value(s) where you want lines
      add = TRUE, lwd = 2, col = "black",
      drawlabels = TRUE,
      method = "flattest",
      vfont = c("sans serif", "bold"),
  labcex = 0.8                      # optional: bigger labels
    )
  },
  key.title = title(main = bquote(atop("log"[10], "peak flux")), cex.main = 1)
)
dev.off()

```

```{r parms-peak-decay-flux}
# generation of new response
Ap = 100 #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = 1 # initial specific ab concentration
tp = 30 #time of peak titer -- days

# steady state nonspecific antibodies
t_halfs = c(30,90,400,4000) #half lives of mucosal ab in days - 1 month, 3 months, 1 year, 10 years :) 
d_mus = log(2)/t_halfs #half lives in days
  names(d_mus) = t_halfs
c = Ap*(log(2)/min(t_halfs))# calculate c such that the minimum M_ns = Ap = 100
Mns_vec = c/d_mus #vector of approximate ss solutions
  names(Mns_vec) = t_halfs

# transport
Vmax = 10^3 # order of magnitude of peak response # we might vary this later
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
KM = Ap/3 #choosing KM such that at peak in the absence of Mns, Ms would achieve 75% of max flux at peak
k1 = k2/KM #k2/min(Mns_vec) #rate of antibody binding such that KM is on the order of 10

### parms for heatmap 
Aps = seq(100, 1000, by = 10) #peak antibody titer will be the same as Mns for 90 days

t_halfs = 3*10^seq(1,3.1, by = 0.01) #half lives of mucosal ab in days - 1 month, 3 months, 1 year, 10 years :) 
 
heatmap_matrix = sapply(Aps, function(Ap){ #columns
  sapply(t_halfs, function(t_half){ #rows
    #set other parms
    d_mu = log(2)/t_half #half lives in days
    Mns = c/d_mu #vector of approximate ss solutions
    
    #calculate peak flux
    phi_s = MM_function(M_s = Ap, M_ns = Mns)
  })
})

### plot
#plot AICs
filled.contour(x = log10(t_halfs), y = (Aps), z=log10(heatmap_matrix), lwd = 2, ylab = expression(paste("peak response")), xlab = expression(paste(log[10], " exponential half life, ", days)), main = bquote("log"[10] ~ "Peak Flux of Specific Antibody"))


## compute the colour-band breaks explicitly, on the log10 scale
### plot on log10 scale
x_ticks = c(30, 100, 300, 1000, 3000)   # half-life ticks (days)
y_ticks = seq(100,1000, by = 100)    # peak titer ticks
zlog <- log10(heatmap_matrix)
lev  <- pretty(range(zlog, finite = TRUE), 5)   # 20 = filled.contour's default nlevels

png(here("test.png"), width = 2000, height = 1200, res = 300, type = "cairo", antialias = "none")
filled.contour(
  x = log10(t_halfs),
  y = Aps,
  z = zlog,
  levels = lev,
  xlab = expression(paste("Plasma cell half life, ", days)),
  ylab = "Peak specific antibody response in mucosa",
  main = "Peak Flux of Specific Antibody into Lumen",
  plot.axes = {
    axis(1, at = log10(x_ticks), labels = x_ticks)
    axis(2, at = seq(200, 1000, by = 200), labels = seq(200, 1000, by = 200), cex.axis = 0.8)
    contour(
      x = log10(t_halfs), y = Aps, z = zlog,
      levels = c(0.5,1,1.5,2,2.5,3),
      add = TRUE, lwd = 2, col = "black",
      drawlabels = TRUE,
      method = "edge",
      vfont = c("sans serif", "bold"),
  labcex = 0.8                      # optional: bigger labels
    )
  },
  key.title = title(main = bquote(atop("log"[10], "peak flux")), cex.main = 1)
)
dev.off()
```

# Part 2 - Inducible Plasma Cells

Below, we define an ode model in which there are 3 populations of cells: B cells, secretory plasma cells, and quiescent plasma cells. We include a fourth variable - "Virus" - which is either 0 (no infection present) or 1 (infection present). 

When $V = 0$, each cell population decays at some rate, and secretory plasma cells differentiate into quiescent plasma cells. 

When $V = 1$, B cells proliferate and differentiate into secretory plasma cells. Secretory plasma cells differentiate into quiescent plasma cells at some rate. Quiescent plasma cells are recruited to differentiate into secretory plasma cells at some rate. Cell populations continue to decay according to their given death rate. 

```{r define-ODE-Model-1}
# Antibody concentrations in the respiratory mucosa #
secondary_response.model1 <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        B = y[1]         # B cells 
        P_s = y[2]       # Antibody secreting cells (specific)
        P_q = y[3]       # quiescent antibody secreting cells (specific)
        M_s = y[4]       # specific Ab in mucosa
        
        # make empty variables for the derivatives
        dB = 0      # change in specific Ab in the lumen
        dP_s = 0    # change in specific ASCs
        dP_q = 0    # change in quiescent ASCs
        dM_s = 0    # change in specific antibodies in mucosa
        
        #####################################
        # define V function 
        V.function = function(t_infection, t) {
          if(t>=min(t_infection) & t <= max(t_infection)){
            V = 1}
          else(V = 0)
        return(V)}
        # set V
        V = V.function(t_infection, t)
        #####################################
        ## calculate the derivatives
        # cell populations
        dB = r*(1-f)*V*B - d_B*B #proliferation only when virus present, some fraction f differentiates to P_s
        dP_s = r*f*V*B + s*V*P_q - d_s*P_s - q*P_s #differentiation from B cells or P_q only when virus present
        dP_q = q*P_s - d_q*P_q - s*V*P_q #P_s always differentiates to P_q with some rate q.
        
        # Specific antibodies        
        dM_s = a*P_s - d_a*M_s #P_s secretes antibody
        
        # output the derivatives
        dy=c(dB, dP_s, dP_q, dM_s) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} # we will need to think about how to set M_ns if we want to track specific IgA in the lumen.

```

```{r set-parms-ODE-Model-1}
# death rates
d_B = log(2)/(365) # B cell death rate, x days half life
d_s = 0 # secretory cell death rate #don't consider for now
d_q = log(2)/(365) # quiescent cell death rate long lived like bm
  
# proliferation and differentiation
r = 0.3 # b cell proliferation rate - doubling time is 1/2 day
f = 1/3 # fraction of division that produce secretory cells
q = log(2)/14 # rate of secretory cells to quiescence #should equal d_mu
s = 10 # rate of quiescent cell recruitment to secretion #should be very fast, like a percentage per hour

# antibody dynamics
a = 100 # rate of IgA secretion by secretory cells # not dealing with density but total number now
d_a = log(2)/3 # IgA molecular decay rate, 3 days half life

# infection timing
t_infection = c(0,14) #time range of antigen presence (V = 1 when t in t_infection)
```

```{r solve-ODE-Model-1-infection}
# parameters list for deSolve
params = list(
  d_B = d_B, # B cell death rate
  d_s = d_s, # secretory cell death rate
  d_q = d_q, # quiescent cell death rate
  d_a = d_a, # IgA molecular decay rate
    
  # proliferation and differentiation
  r = r, # b cell proliferation rate
  f = f, # fraction of division that produce secretory cells
  q = q, # rate of secretory cells to quiescence
  s = s, # rate of quiescent cell recruitment to secretion
  
  # antibody dynamics
  a = a, # rate of IgA secretion by secretory cells
  d_a = d_a, # IgA molecular decay rate in mucosa
  
  #infection timing
  t_infection = t_infection #time range of antigen presence (V = 1 when t in t_infection)
)

# simulation Time parameters 
Tmax   = 30                         # max time in days
points.per.day = 24              # output points per day

#initial conditions
init_cond = c("B" = 3, "P_s" = 0, "P_q" = 1, "M_s" = 0)

# solve 
soln = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model1, parms = params)) 

# long form ode 
soln_long = soln %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
   coord_cartesian(xlim = c(0,5), ylim = c(0,5)) +
   labs(x = "days",, y = bquote("log"[10] ~ "number"), title = "primary infection, inducible response") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)
```

We can compare this to a nearly identical model, where secretory plasma cells die rather than becoming quiescent and this death occurs at the same rate as quiescence: 

compare to if $d_\mu$ was actual death rate

```{r ODE-model-no-inducible}
# Antibody concentrations in the respiratory mucosa #
secondary_response.model_notinducible <- function(t, y, parms)  # Single partial immune class (RPS)                       
{with(as.list(parms),	# allows the parameter file parms to be as a list
      {  
        y = pmax(y,0)         # avoid negative values (not ideal but ok)
        
        # map the state variables
        B = y[1]         # B cells 
        P_s = y[2]       # Antibody secreting cells (specific)
        M_s = y[3]       # specific Ab in mucosa
        
        # make empty variables for the derivatives
        dB = 0      # change in specific Ab in the lumen
        dP_s = 0    # change in specific ASCs
        dM_s = 0    # change in specific antibodies in mucosa
        
        #####################################
        # define V function 
        V.function = function(t_infection, t) {
          if(t>=min(t_infection) & t <= max(t_infection)){
            V = 1}
          else(V = 0)
        return(V)}
        # set V
        V = V.function(t_infection, t)
        #####################################
        ## calculate the derivatives
        # cell populations
        dB = r*(1-f)*V*B - d_B*B #proliferation only when virus present, some fraction f differentiates to P_s
        dP_s = r*f*V*B - d_s*P_s #differentiation from B cells or P_q only when virus present
        
        # Specific antibodies        
        dM_s = a*P_s - d_a*M_s #P_s secretes antibody
        
        # output the derivatives
        dy=c(dB, dP_s, dM_s) #vector of derivatives 
        
        return(list(dy)) #lists vector and returns
      }
  ) 
} # we will need to think about how to set M_ns if we want to track specific IgA in the lumen.
```

```{r set-parms-ODE-Model-1-noinducible}
# death rates
d_B = d_B # B cell death rate, 2 years half life
d_s = q # rate of secretory cells to death #should equal d_mu and q
 
# proliferation and differentiation
r = r # b cell proliferation rate - doubling time is 1 day
f = f # fraction of division that produce secretory cells

# antibody dynamics
a = a # rate of IgA secretion by secretory cells # not dealing with density but total number now
d_a = d_a # IgA molecular decay rate, 3 days half life

# infection timing
t_infection = c(0,14) #time range of antigen presence (V = 1 when t in t_infection)
```

```{r solve-ODE-Model-1-infection-noinducible}
# parameters list for deSolve
params = list(
  d_B = d_B, # B cell death rate
  d_s = d_s, # secretory cell death rate

    
  # proliferation and differentiation
  r = r, # b cell proliferation rate
  f = f, # fraction of division that produce secretory cells

  # antibody dynamics
  a = a, # rate of IgA secretion by secretory cells
  d_a = d_a, # IgA molecular decay rate in mucosa
  
  #infection timing
  t_infection = t_infection #time range of antigen presence (V = 1 when t in t_infection)
)

# simulation Time parameters 
Tmax   = 30                         # max time in days
points.per.day = 24              # output points per day

#initial conditions
init_cond = c("B" = 3, "P_s" = 0, "M_s" = 0)

# solve 
soln_notinducible = as.data.frame(ode(y = init_cond, times = seq(0, Tmax, length.out = Tmax*points.per.day), secondary_response.model_notinducible, parms = params)) 

# long form ode 
soln_notinducible_long = soln_notinducible %>% pivot_longer(
  cols = !c(time),
  names_to = "population",
  values_to = "number"
)

# ggplot
 plot = ggplot() +
  geom_point(data = soln_notinducible_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time 
  geom_line(data = soln_notinducible_long, aes(x = time, y = log10(number), group = population, col = population)) + #plots each population seperately over time
   coord_cartesian(ylim = c(0,5), xlim = c(0,5)) +
   labs(x = "days", y = bquote("log"[10] ~ "number"), title = "primary infection, not inducible response") +
  theme_bw() + #removes grey background and changes aesthetics
  scale_color_manual(values = my_colors) #changes colors to defined vector

# print
print(plot)

```

To compare the two, we plot the changes in $P_s$ and $M_s$ for both models over the first three daysL 

```{r plot-together}
plot = ggplot() + geom_line(data = soln_long %>% filter(population %in% c("P_s", "M_s")), aes(x = time, y = (number), linetype = "inducible", group = population, color = population), lwd = 1.5) + 
  geom_line(data = soln_notinducible_long %>% filter(population %in% c("P_s", "M_s")), aes(x = time, y = (number), linetype = "not inducible", group = population, color = population), lwd = 1.5) +
  coord_cartesian(xlim = c(0,30), ylim = 10^c(0,3)) +
  scale_linetype_manual(values = c("inducible" = "solid", "not inducible" = "dashed")) +
  labs(x = "days", y = bquote("Ms"), linetype = "model", title = "secondary infection, antibody response") +
  theme_bw() + #removes grey background and changes aesthetics
  theme(axis.text = element_text(size = 25),
        axis.title = element_text(size = 25),
        legend.title = element_text(size = 25),
        legend.text = element_text(size = 25, hjust = 0),
        legend.key.width = unit(0.06, "npc")
  )
#print to viewer
print(plot)


```