---
title: "Constraints on the longevity of IgA: Simulations of specific IgA transport in the context of pre-existing, nonspecific immunity"
subtitle: "Paper Figures"
format: html
---

```{r setup}
#install.package("here")
library(here)

#install.package("ggplot2")
library(ggplot2)

#install.package("gridExtra")
library(gridExtra)

#install.packages("tidyverse")
library(tidyverse)

#install.packages("deSolve")
library(deSolve)

#load functions
MM_function = function(M_s, M_ns) {(Vmax*M_s)/(KM+M_ns+M_s)}


#choose plotting colors
my_colors = c("darkred", "maroon", "darkblue", "forestgreen", "darkorange", "goldenrod") #choose color range
num_colors = length(my_colors) #set number of colors needed for plotting
my_colors = colorRampPalette(my_colors)(num_colors) #increase number of colors staying in color range
```

# Objective 

In part 1 this document, we evaluate how long-lived IgA antibody titers in the respiratory mucosa interfere with the transport of specific IgA antibodies to the site of infection. We assume that the concentration of $M_{ns}$ changes negligibly over the course of infection, and assume that the steady state concentration of $M_{ns}$ is proportional to $\frac{c}{d_\mu}$. We use a Michaelis-Mentin approximation to estimate the flux of specific IgA across the mucosal membrane. 


# Outline 

I will review the kinetics inferred by the Michaelis-Mentin approximation before presenting the results of competitive inhibition by non-specific immunity during a specific response. 

The results in this document are organized as follows: 

1) First, we will explore the effect of changing $d_\mu$ on the total flux of specific IgA, $k_2R_S$ (or $\phi_S$) over time. We will demonstrate how much specific IgA must be generated during a response to overcome the effect of competition from pre-existing titers. 



# Competition for transport: Michaelis-Mentin Kinetics

Below, we define our parameters: 

```{r define-parameters-MM}
# generation of new response
Ap = 100 #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = 1 # initial specific ab concentration
tp = 30 #time of peak titer -- days

# steady state nonspecific antibodies
t_halfs = c(30,90,400,4000) #half lives of mucosal ab in days - 1 month, 3 months, 1 year, 10 years :) 
d_mus = log(2)/t_halfs #half lives in days
  names(d_mus) = t_halfs
c = Ap*(log(2)/min(t_halfs))# calculate c such that the minimum M_ns = Ap = 100
Mns_vec = c/d_mus #vector of approximate ss solutions
  names(Mns_vec) = t_halfs


# transport
Vmax = 10^3 # order of magnitude of peak response # we might vary this later
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
KM = Ap/3 #choosing KM such that at peak in the absence of Mns, Ms would achieve 75% of max flux at peak
k1 = k2/KM #k2/min(Mns_vec) #rate of antibody binding such that KM is on the order of 10

# dynamics in the lumen 
t_half_L = 3 #half life of antibodies in the lumen, in days
d_L = log(2)/t_half_L #decay rate in the lumen


# Total time to plot responses
Tmax = 90

# will be calculated when iterating through d_mu
#r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
#alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate


```


Below, we plot how the flux of specific antibody for a given concentration of $M_s$ in the mucosa changes as a function of the half-life of non-specific antibodies. In essence, the effective amount of $M_s$ required to reach half-maximal transport will increase by the concnetration of non-specific antibody conferred by the increasing half-life. 

```{r plot-MM-kinetics}
#set concentrations of Ms for x axis
concentrations = 10^seq(-1, 4, by = 0.1) #range of concentrations M_s such that we span much less than and much greater than Ap

### plot flux vs [Ms] given each Mns 
flux_df = lapply(t_halfs, function(t_half){
  # set d_mu and Mns 
  d_mu = log(2)/t_half
  Mns = c/d_mu
  
  #calculate flux
  flux = sapply(concentrations, function(Ms){
   
   return(MM_function(M_s = Ms, M_ns = Mns))
  }) %>% cbind("Ms" = concentrations, "flux" = ., "t_half" = t_half) %>% as.data.frame() 
}) %>% do.call(rbind, .) %>% as.data.frame() 

## plot
plot = ggplot() + geom_line(data = flux_df, aes(x = (Ms), y = flux, group = factor(t_half), col = factor(t_half))) +
  geom_hline(yintercept = Vmax, linetype = "dashed") +
  geom_vline(xintercept = (KM), color = "red") +
  geom_vline(xintercept = (Ap), linetype= "dashed") +
geom_line(
  aes(x = concentrations, y = sapply(concentrations, function(Ms){Vmax*Ms/(KM+Ms)})),
  inherit.aes = FALSE,
  color = "red"
) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = my_colors) +
  labs(x = "[Ms]", y = "flux [Ms]", color = "Mucosal Ab Half Life, days", title = "Competitive inhibition of specific antibody transport by nonspecific antibodies") + 
  theme_bw()

# print to viewer
print(plot)
```

The horizontal dashed line shows the greatest possible transport rate, $V_{max}$; the red curve shows the MM kinetics given $[M_s]$ only; the vertical red line shows the 50% saturation concentration given $[M_s]$ only, $K_M$; and I have added a vertical dashed line to show the peak concentration of $M_s$ generated over an infection.

Notice that I have chosen paramerters such that, in the absence of competition, the maximum flux is reached by the the maximum concentration of $M_s$ generated during a response. As nonspecific antibody is added, however, the flux achieved at the maximum specific antibody concentration drops. This is the core of the competitive effect. 

Below, I show the generation and decay of $M_s$ over the course of infection with respect to the steady-state concentration of $M_{ns}$ for increasing half-life of mucosal antibodies. 

```{r plot-Ms-and-Mns}
### plot Ms vs Mns over time
Mab_titers = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = c/d_mu # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  # set time vector parms
  Tmax = Tmax # total days
  points_per_day = 24 # every hour
  time = seq(0, Tmax, length.out = Tmax*points_per_day) #generate time vector

  ### apply to calculate
  soln = data.frame("time" = time, "Ms" = sapply(time, function(t){
    if(t<=tp){A0*exp((r0 - d_mu)*t - (alpha/2)*t^2)}
    else {Ap*exp(-d_mu*(t-tp))}
  }))
  

  # solve 
  soln = soln %>% cbind(.,Mns, t_half)
}) %>% do.call(rbind, .) #concatenates all dfs from lapply list into a single df rowwise

### Plot
plot = ggplot() + geom_line(data = Mab_titers, aes(x = time, y = log10(Ms), group = factor(t_half), col = factor(t_half), linetype = "Ms"), lwd = 1.5) +
  geom_line(data = Mab_titers, aes(x = time, y = log10(Mns), group = factor(t_half), col = factor(t_half), linetype = "Mns"), lwd = 1.5) + 
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(values = c("Ms" = "solid", "Mns" = "dashed")) +
  labs(x = "time, days", y = "log10[Ab], mg/dL", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "[Ms] and [Mns] over the course of infection") +
  theme_bw()
  

#print to viewer
print(plot)

```

I have chosen the peak of the response such that the peak titer matches the steady state of the nonspecific antibody for an intermediate half-life. 

# Results 1: Flux over time as a function of $d_\mu$ or $\frac{k_2}{k_1}$

The flux of specific IgA over time will depend on competition with pre-existing immunity. This competition changes according to i) the amount of non-specific pre-exisitng immunity, which is a function of the rate of decay of responses, and ii) the magnitude of the peak specific antibody titer with respect to the saturation of transport, which we can change as a funtion of $K_M$, the half saturation point, which equals $\frac{k_2}{k_1}$. 

## Flux over time as a function of $d_\mu$

Below, we calculate the flux of specific IgA over time as a function of the decay rate of previously generated IgA: 

```{r parms-flux-decay-rate}
# set parms 
#transport 
Vmax = Vmax
k2 = k2
RT = RT
k1 = k1
KM = KM

#lumen
t_half_L = t_half_L
d_L = d_L

#steady state Ab
c = c
t_halfs = t_halfs
d_mus = d_mus
Mns_vec = Mns_vec
Ap = Ap

# generating new response
Ap = Ap
A0 = A0
tp = tp

# Total time 
Tmax = Tmax
```

```{r calculate-flux-decay-rate}
MM_soln = lapply(t_halfs, function(t_half){
  # setting t.half dependent parms for nonspecific population
  d_mu = log(2)/t_half #set d_mu
  Mns = c/d_mu # steadys state nonspecific
  
  # setting t.half dependent parms for specific abs
  r0 = (2*log(Ap/A0)/tp) + d_mu #initial/greatest exponential expansion rate
  alpha = (r0 - d_mu)/tp # rate of decay of exponential growth rate

  # set time vector parms
  Tmax = Tmax # total days
  points_per_day = 24 # every hour
  time = seq(0, Tmax, length.out = Tmax*points_per_day) #generate time vector

  ### apply to calculate Ms titers over time
  Ms_soln = data.frame("time" = time, "t_half" = t_half, "Ms" = sapply(time, function(t){
    if(t<=tp){A0*exp((r0 - d_mu)*t - (alpha/2)*t^2)}
    else {Ap*exp(-d_mu*(t-tp))}
  }), "Mns" = Mns)
  
  ### apply to calculate Ls flux over time
  Ms_soln$phi_s = sapply(Ms_soln$Ms, function(Ms){MM_function(M_s = Ms, M_ns = Mns)})
  Ms_soln$phi_ns = sapply(Ms_soln$Ms, function(Ms){(Vmax*Mns)/(KM+Ms+Mns)})
  
  return(Ms_soln)
}) %>% do.call(rbind, .)

```

```{r plot-flux-decay-rate}
#ggplot
plot = ggplot() + geom_line(data = MM_soln, aes(x = time, y = (phi_s), group = factor(t_half), col = factor(t_half), linetype = "phi_s"), lwd = 1.5) +
  geom_line(data = MM_soln, aes(x = time, y = (phi_ns), group = factor(t_half), col = factor(t_half), linetype = "phi_ns"), lwd = 1.5) + 
  scale_color_manual(values = my_colors) +
  #coord_cartesian(ylim = c(0, 3), xlim = c(0, 15)) +
  scale_linetype_manual(values = c("phi_s" = "solid", "phi_ns" = "dashed")) +
  labs(x = "time, days", y = "[R-Ab], mg/dL, over time", col = "mucosal Ab half-life, days", linetype = "Specificity", title = "flux of s and ns into the lumen over the course of infection") +
  theme_bw()

print(plot)

```

yay

Now, we would like to plot how the peak flux (or the total amount of antibody transported) changes as a function of $t_{\frac{1}{2}}$ and the magnitude of the response in the mucosa. This way, we can see how the magnitude of the response required increases with $t_{1}{2}$

```{r parms-peak-decay-flux}
# generation of new response
Aps = 10^seq(1, 3, by = 0.01) #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = 1 # initial specific ab concentration
tp = 30 #time of peak titer -- days

# steady state nonspecific antibodies
t_halfs = 3*10^seq(1,3.1, by = 0.01) #half lives of mucosal ab in days - 1 month, 3 months, 1 year, 10 years :) 

 
heatmap_matrix = sapply(Aps, function(Ap){ #columns
  sapply(t_halfs, function(t_half){ #rows
    #set other parms
    d_mu = log(2)/t_half #half lives in days
    c = Ap*(log(2)/min(t_halfs))# calculate c such that the minimum M_ns = Ap = 100
    Mns = c/d_mu #vector of approximate ss solutions
    
    # transport
    Vmax = 10^3 # order of magnitude of peak response # we might vary this later
    k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
    RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
    KM = Ap/3 #choosing KM such that at peak in the absence of Mns, Ms would achieve 75% of max flux at peak
    k1 = k2/KM #k2/min(Mns_vec) #rate of antibody binding such that KM is on the order of 10
    
    # dynamics in the lumen 
    t_half_L = 3 #half life of antibodies in the lumen, in days
    d_L = log(2)/t_half_L #decay rate in the lumen

    # Total time to plot responses
    Tmax = 90
    
    #calculate peak flux
    phi_s = MM_function(M_s = Ap, M_ns = Mns)
  })
})

### plot
### plot on log10 scale
x_ticks = c(30, 100, 300, 1000, 3000)   # half-life ticks (days)
y_ticks = c(10, 30, 100, 300, 1000)     # peak titer ticks
#plot AICs
filled.contour(x = log10(t_halfs), y = log10(Aps), z=heatmap_matrix, lwd = 2, ylab = expression(paste("peak specific response")), xlab = expression(paste(log[10], " exponential half life, ", days^{-1})), main = "Peak Flux of Specific Antibody")

filled.contour(
  x = log10(t_halfs),
  y = log10(Aps),
  z = heatmap_matrix,
  xlab = "exponential half life (days)",
  ylab = "peak antibody titer",
  main = "Peak Flux of Specific Antibody",
  plot.axes = {
    axis(1, at = log10(x_ticks), labels = x_ticks)
    axis(2, at = log10(y_ticks), labels = y_ticks)
    contour(
      x = log10(t_halfs),
      y = log10(Aps),
      z = heatmap_matrix,
      levels = seq(100,400, by = 100),   # the flux value(s) where you want lines
      add = TRUE,
      lwd = 2,
      col = "black",
      drawlabels = TRUE
    )
  },
  key.title = title(main = "peak flux", cex.main = 1)
)

```

```{r parms-peak-decay-flux}
# generation of new response
Ap = 100 #peak antibody titer will be the same as Mns for 90 days
#initial ab titer
A0 = 1 # initial specific ab concentration
tp = 30 #time of peak titer -- days

# steady state nonspecific antibodies
t_halfs = c(30,90,400,4000) #half lives of mucosal ab in days - 1 month, 3 months, 1 year, 10 years :) 
d_mus = log(2)/t_halfs #half lives in days
  names(d_mus) = t_halfs
c = Ap*(log(2)/min(t_halfs))# calculate c such that the minimum M_ns = Ap = 100
Mns_vec = c/d_mus #vector of approximate ss solutions
  names(Mns_vec) = t_halfs

# transport
Vmax = 10^3 # order of magnitude of peak response # we might vary this later
k2 = 1/(2/24) #rate of transport - 1/T of transport which is 1/2hours or 1/2/24ths of a day
RT = Vmax/k2 #total number (concentration) of receptors, Vmax = k2RT
KM = Ap/3 #choosing KM such that at peak in the absence of Mns, Ms would achieve 75% of max flux at peak
k1 = k2/KM #k2/min(Mns_vec) #rate of antibody binding such that KM is on the order of 10

### parms for heatmap 
Aps = 10^seq(1, 3, by = 0.01) #peak antibody titer will be the same as Mns for 90 days

t_halfs = 3*10^seq(1,3.1, by = 0.01) #half lives of mucosal ab in days - 1 month, 3 months, 1 year, 10 years :) 
 
heatmap_matrix = sapply(Aps, function(Ap){ #columns
  sapply(t_halfs, function(t_half){ #rows
    #set other parms
    d_mu = log(2)/t_half #half lives in days
    Mns = c/d_mu #vector of approximate ss solutions
    
    #calculate peak flux
    phi_s = MM_function(M_s = Ap, M_ns = Mns)
  })
})

### plot
#plot AICs
filled.contour(x = log10(t_halfs), y = log10(Aps), z=heatmap_matrix, lwd = 2, ylab = expression(paste("peak response")), xlab = expression(paste(log[10], " exponential half life, ", days)), main = "Peak Flux of Specific Antibody")



```

